---
title: "Combine_adnimerge_TA_data_allADNI"
author: "Alfie Wearn"
date: "25/03/2022"
output: pdf_document
---
```{r}
# This will add Texture Analysis results to ADNIMERGE data sheet. 
# Result will be the ability to plot and analyse the ADNI data 
# in relation to TA to your heart's content. However, ADNI1 data should not be excluded at this point.
# ppts with bad ASHS data should not be given ASHS data, but should not be removed entirely (other variables should remain)

# Written by Alfie Wearn 2021-07-21.
library(tidyverse)

# Identify the data directories
adni_dir <- "/Users/alfiewearn/MRIdata/ADNI/Clinical_Data"
ashs_dir <- "/Users/alfiewearn/MRIdata/ADNI/ASHS_output/T1_only"
ta_data_dir <- "../data" # Assumes R project is in ADNI/texture_analysis_processing/code
```

Will you exclude bad ASHS participants? 
```{r}
exclude_bad_ashs = TRUE
```


# Read in data files
```{r}
adnimerge <- read.csv(paste(adni_dir,"/../adnimerge_update/Mar2022/ADNIMERGE.csv", sep=""))
av1451_data <- read.csv(paste(adni_dir,"/UCBERKELEYAV1451_01_14_21.csv", sep="")) # AV45 data
# av45_data <- read.csv(paste(adni_dir,"/UCBERKELEYAV45_01_14_21.csv", sep="")) # AV45 data
# fbb_data <- read.csv(paste(adni_dir,"/UCBERKELEYFBB_01_14_21.csv", sep="")) # FBB
```


```{r}
ta_data <- read.csv(paste(ta_data_dir, "/ta_stats_z_denoise.csv", sep=""))
```

# Remove an erroneous row from ADNIMERGE
RID 6014 has both 'bl' and 'm0'. m0 is identical to 'bl' but is not the correct format, and appears to just be a duplicate. Delete this row.
```{r}
adnimerge <- adnimerge %>%
  filter(!(RID=='6014' & VISCODE=='m0'))
```


Exclude bad ashs
```{r}
if (exclude_bad_ashs==TRUE) {
  exclusions <- read.csv(paste(ashs_dir,"/bad_ASHS.csv", sep=""), header=FALSE)
  
  # Give column names to exclusions file
  colnames(exclusions) <- c("PTID","VISCODE")
  
  # Delete volume and TA data for people and sessions in exclusions (keep Region and Bins columns though)
  bad_data <- ta_data %>%
    right_join(exclusions, by = c("PTID","VISCODE")) 
  bad_data[,5:length(bad_data)] <- NA
  
  ta_data_good <- ta_data %>%
    anti_join(bad_data, by = c("PTID","VISCODE","Region","bins")) %>%
    rbind(bad_data) %>%
    distinct
  
  merged_data <- merge(adnimerge, ta_data_good,  by = c("PTID", "VISCODE"), all.x = TRUE) %>% distinct()
} else {
  merged_data <- merge(adnimerge, ta_data, by = c("PTID", "VISCODE"))  %>% distinct()
}
```


```{r}
# Remove non numeric characters from certain columns
merged_data$TAU <- str_replace_all(merged_data$TAU, "<", "") %>% as.double()
merged_data$PTAU <- str_replace_all(merged_data$PTAU, "<", "") %>% as.double()
merged_data$ABETA <- str_replace_all(merged_data$ABETA, ">", "") %>% as.double()

# Change glrlm_LRLGLE to a double, not a factor (why is it even a factor...?)
merged_data$glrlm_LRLGLE <- merged_data$glrlm_LRLGLE %>% as.character() %>% as.double() 


# CHANGE DATA TO NUMERIC
first_TA_col_md <- which(colnames(merged_data)=="Volume") %>% as.integer()
last_TA_col_md <- which(colnames(merged_data)=="glrlm_LRLGLE") %>% as.integer()

# change all TA columns to numeric.
merged_data[,first_TA_col_md:last_TA_col_md] <- sapply(merged_data[,first_TA_col_md:last_TA_col_md],as.numeric) %>% as.data.frame()
```


# Merge in Florbetapir and Florbetapen (former just to compare, latter for calculation of centiloids)
This does not need to be done if using ADNIMERGE from March2022 (or later), as FBB is already included!
```{r, eval = FALSE}
av45_data_to_merge <- av45_data %>% select(RID, VISCODE2, SUMMARYSUVR_WHOLECEREBNORM) %>% rename(VISCODE = VISCODE2)
new_merge <- merge(av45_data_to_merge, merged_data, by=c("RID","VISCODE"), all.y=TRUE) %>% rename(AV45_raw = SUMMARYSUVR_WHOLECEREBNORM)

fbb_data_to_merge <- fbb_data %>% select(RID, VISCODE2, SUMMARYSUVR_WHOLECEREBNORM) %>% rename(VISCODE = VISCODE2)
new_merge_fbb <- merge(fbb_data_to_merge, new_merge, by=c("RID","VISCODE"), all.y=TRUE) %>% rename(FBB_raw = SUMMARYSUVR_WHOLECEREBNORM)

merged_data <- new_merge_fbb %>% relocate(AV45_raw, FBB_raw, .after=AV45)
```

Get AV1451 info (Flortaucipir) - Tau PET.
```{r}
av1451_data_to_merge <- av1451_data %>% select(RID, VISCODE2, META_TEMPORAL_SUVR) %>% rename(VISCODE = VISCODE2)
new_merge <- merge(av1451_data_to_merge, merged_data, by=c("RID","VISCODE"), all.y=TRUE) %>% rename(AV1451 = META_TEMPORAL_SUVR)

merged_data <- new_merge %>% relocate(AV1451, .after=FBB)
```


# SAVE THE DATA 
```{r}
write.csv(merged_data, file = "../data/merged_data.csv",row.names = FALSE)
saveRDS(merged_data, file = "../data/merged_data.Rdata")
```


```{r}
system("say Data Merged")
```


