---
title: "Create_models"
author: "Alfie Wearn"
date: "10/03/2022"
output: pdf_document
---

Based on Lars' send code 'Mulvar_Modelling_LLR', this script uses splines to model trajectory of texture variables

Load packages and data
```{r}
library(progmod)
library(tidyverse)
library(R.utils) # For 'insert' function

spline_mean <- function(t, s, a1, a2, a3, a4, a5) {
  a <- c(a1[1], a1[1], a2[1], a3[1], a4[1], a5[1], a5[1])
  mf <- splinefun(x = c(-101, seq(-100, 200, length = 5), 201),  # Anchor points should be chosen to match the data range
                  y = a, 
                  method = 'natural')
  
  mf(t + s)
}
```
Load data
```{r}
load(file="../data/prepped_data.Rdata")
data <- selected_data
```
 

Get sample size info
```{r}
# All - 3 groups
data %>%
  filter(RID != 4386) %>% # The one with absurd values for AV45...
  filter(!is.na(CDRSB) | !is.na(MMSE) | !is.na(ADAS13) | !is.na(AB_centiloid)) %>% #  Only one of each variable required
  group_by(RID) %>%
  filter(M == min(M)) %>% # pick one session for each subject to count total subjects
  ungroup() %>%
  group_by(DX_baseline) %>%
  summarise(n=n(),
            Mage = mean(Age_bl),
            SDage = sd(Age_bl))

# All with amyloid - 3 groups
data %>%
  filter(RID != 4386) %>% # The one with absurd values for AV45...
  filter(!is.na(CDRSB) | !is.na(MMSE) | !is.na(ADAS13) | !is.na(AB_centiloid)) %>% #  Only one of each variable required
  filter(!is.na(DX_baseline_Ab)) %>%
  group_by(RID) %>%
  filter(M == min(M)) %>% # pick one session for each subject to count total subjects
  ungroup() %>%
  group_by(DX_baseline) %>%
  summarise(n=n(),
            Mage = mean(Age_bl),
            SDage = sd(Age_bl))

# All with amyloid - 4 groups
data %>%
  filter(RID != 4386) %>% # The one with absurd values for AV45...
  filter(!is.na(CDRSB) | !is.na(MMSE) | !is.na(ADAS13) | !is.na(AB_centiloid)) %>% #  Only one of each variable required
  group_by(RID) %>%
  filter(M == min(M)) %>% # pick one session for each subject to count total subjects
  ungroup() %>%
  group_by(DX_baseline_Ab) %>%
  summarise(n=n(),
            Mage = mean(Age_bl),
            SDage = sd(Age_bl))

# How many sessions? 2356
data %>%
  filter(RID != 4386) %>% # The one with absurd values for AV45...
  filter(!is.na(CDRSB) | !is.na(MMSE) | !is.na(ADAS13) | !is.na(AB_centiloid)) %>% #  Only one of each variable required
  nrow()

# Save ID list
model_train_list <- data %>%
  filter(RID != 4386) %>% # The one with absurd values for AV45...
  filter(!is.na(CDRSB) | !is.na(MMSE) | !is.na(ADAS13) | !is.na(AB_centiloid)) %>% #  Only one of each variable required
  filter(!is.na(DX_baseline_Ab)) %>%
  select(PTID, VISCODE) %>%
  mutate(model_train_subset = "Y")

save(model_train_list, file="../data/ID_lists/model_train_list.Rdata")
```


Make long dataset with relevant variables
```{r}
prog_dat <- data[, c('RID', 
                     'Age_bl',
                     'male',
                     'PRE', 'MCI', 'DEM',  
                     'Month_bl',
                     'MMSE', 'CDRSB', 'ADAS13', 'AB_centiloid')]

# Remove v high AV45 value
prog_dat <- prog_dat %>%
  filter(RID != 4386) # The one with absurd values for AV45...
hist(prog_dat$AB_centiloid)

prog_dat <- pivot_longer(prog_dat, cols = c('MMSE', 'CDRSB', 'ADAS13','AB_centiloid'))
prog_dat <- na.omit(prog_dat)
prog_dat$name <- factor(prog_dat$name)
```


Fit univariate model
```{r}
start_coef <- c(l = 0.14, 
                s.PRE = 39, s.MCI = 123, s.DEM = 178,
                g = 3.6, 
                v = 10)

mod_adas_yr <- progmod(value ~ exp_model(Month_bl, l, s, g, v),
                       data = subset(prog_dat, name == 'ADAS13'),
                       fixed = list(l ~ 1,
                                    s ~ PRE + MCI + DEM + 0,
                                    g ~ 1,
                                    v ~ 1),
                       random = s + v ~ 1 | RID,
                       start = start_coef)

AIC(mod_adas_yr) # Using continuous time is best
fixed.effects(mod_adas_yr)
```


Fit joint multivariate model on clinical scales
```{r}
start_coef <- c(l.nameADAS13 = 0.14, l.nameCDRSB = 0.019, l.nameMMSE = -0.01, 
                s.PRE = 39, s.MCI = 123, s.DEM = 178, s.Age_bl = 1,
                g.nameADAS13 = 3.8, g.nameCDRSB = 3.7, g.nameMMSE = 3.5, 
                v.nameADAS13 = 10.2, v.nameCDRSB = 0.29, v.nameMMSE = 29)

random_start <- random.effects(mod_adas_yr)
names(random_start)[2] <- 'v.nameADAS13'
random_start$`v.nameCDRSB` <- 0
random_start$`v.nameMMSE` <- 0

mod_joint_cog <- nlme(value ~ exp_model(Month_bl, l, s, g, v),
                      data = subset(prog_dat, name %in% c('MMSE', 'CDRSB', 'ADAS13')),
                      fixed = list(l ~ name + 0,
                                   s ~ PRE + MCI + DEM + Age_bl + 0,
                                   g ~ name + 0,
                                   v ~ name + 0),
                      random = list(s ~ 1,
                                    v ~ name + 0),
                      groups = ~ RID,
                      start = list(fixed = start_coef,
                                   random = random_start),
                      weights = varIdent(form = ~ 1 | name),
                      control = nlmeControl(pnlsMaxIter = 100,
                                            niterEM = 100,
                                            msMaxIter = 100))
fixed.effects(mod_joint_cog)
```


# Spline model
```{r}
fixed_start_coef <- c(a1_adas = 0, a2_adas = 12, a3_adas = 12, a4_adas = 30, a5_adas = 73, 
                      a1_cdrsb = 0, a2_cdrsb = 0.64, a3_cdrsb = 1, a4_cdrsb = 4.8, a5_cdrsb = 18, 
                      a1_mmse = 30, a2_mmse = 27, a3_mmse = 25, a4_mmse = 23, a5_mmse = 4.7, 
                      s.PRE = 37, s.MCI = 102, s.DEM = 164, s.Age_bl = 0)

random_start <- random.effects(mod_joint_cog)
names(random_start)[2:4] <- c('k.nameADAS13', 'k.nameCDRSB', 'k.nameMMSE')
random_start[, 2:4] <- 0

mod_joint_cog_spline <- nlme(value ~ as.numeric(name == 'ADAS13') * spline_mean(Month_bl, s, a1_adas, a2_adas, a3_adas, a4_adas, a5_adas) +
                               as.numeric(name == 'CDRSB') * spline_mean(Month_bl, s, a1_cdrsb, a2_cdrsb, a3_cdrsb, a4_cdrsb, a5_cdrsb) +
                               as.numeric(name == 'MMSE') * spline_mean(Month_bl, s, a1_mmse, a2_mmse, a3_mmse, a4_mmse, a5_mmse) + k,
                             data = subset(prog_dat, name %in% c('ADAS13', 'CDRSB', 'MMSE')),
                             fixed = list(a1_adas + a2_adas + a3_adas + a4_adas + a5_adas ~ 1,
                                          a1_cdrsb + a2_cdrsb + a3_cdrsb + a4_cdrsb + a5_cdrsb ~ 1,
                                          a1_mmse + a2_mmse + a3_mmse + a4_mmse + a5_mmse ~ 1,
                                          s ~ PRE + MCI + DEM + Age_bl + 0),
                             random = list(s ~ 1,
                                           k ~ name + 0),
                             groups = ~ RID,
                             start = list(fixed = fixed_start_coef,
                                          random = random_start
                             ),
                             weights = varIdent(form = ~ 1 | name),
                             control = nlmeControl(msMaxIter = 100))

# system("say spline model done")

AIC(mod_joint_cog) # 82860.49
AIC(mod_joint_cog_spline) # 82589.59

fixed.effects(mod_joint_cog_spline)
```

```{r}
fixed_start_coef <- c(a1_adas = 0, a2_adas = 12, a3_adas = 12, a4_adas = 30, a5_adas = 73, 
                      a1_cdrsb = 0, a2_cdrsb = 0.64, a3_cdrsb = 1, a4_cdrsb = 4.8, a5_cdrsb = 18, 
                      a1_mmse = 30, a2_mmse = 27, a3_mmse = 25, a4_mmse = 23, a5_mmse = 4.7, 
                      s.PRE = 37, s.MCI = 102, s.DEM = 164, s.Age_bl = 0, k.Age_bl = 0)

random_start <- random.effects(mod_joint_cog)
names(random_start)[2:4] <- c('k.nameADAS13', 'k.nameCDRSB', 'k.nameMMSE')
random_start[, 2:4] <- 0

mod_joint_cog_spline_AgeK <- nlme(value ~ as.numeric(name == 'ADAS13') * spline_mean(Month_bl, s, a1_adas, a2_adas, a3_adas, a4_adas, a5_adas) +
                               as.numeric(name == 'CDRSB') * spline_mean(Month_bl, s, a1_cdrsb, a2_cdrsb, a3_cdrsb, a4_cdrsb, a5_cdrsb) +
                               as.numeric(name == 'MMSE') * spline_mean(Month_bl, s, a1_mmse, a2_mmse, a3_mmse, a4_mmse, a5_mmse) + k,
                             data = subset(prog_dat, name %in% c('ADAS13', 'CDRSB', 'MMSE')),
                             fixed = list(a1_adas + a2_adas + a3_adas + a4_adas + a5_adas ~ 1,
                                          a1_cdrsb + a2_cdrsb + a3_cdrsb + a4_cdrsb + a5_cdrsb ~ 1,
                                          a1_mmse + a2_mmse + a3_mmse + a4_mmse + a5_mmse ~ 1,
                                          s ~ PRE + MCI + DEM + Age_bl + 0,
                                          k ~ Age_bl + 0),
                             random = list(s ~ 1,
                                           k ~ name + 0),
                             groups = ~ RID,
                             start = list(fixed = fixed_start_coef,
                                          random = random_start
                             ),
                             weights = varIdent(form = ~ 1 | name),
                             control = nlmeControl(msMaxIter = 100))

# system("say spline model done")

fixed.effects(mod_joint_cog_spline_AgeK)
```


Add predicted disease times
```{r}

AIC(mod_joint_cog,mod_joint_cog_spline,mod_joint_cog_spline_AgeK) 

prog_dat$disease_time_cog <- with(prog_dat, Month_bl + 
                                    fixed.effects(mod_joint_cog)['s.PRE'] * PRE +
                                    fixed.effects(mod_joint_cog)['s.MCI'] * MCI +
                                    fixed.effects(mod_joint_cog)['s.DEM'] * DEM +
                                    fixed.effects(mod_joint_cog)['s.Age_bl'] * Age_bl +
                                    random.effects(mod_joint_cog)[as.character(RID), 's.(Intercept)'])

prog_dat$disease_time_cog_spline <- with(prog_dat, Month_bl + 
                                           fixed.effects(mod_joint_cog_spline)['s.PRE'] * PRE +
                                           fixed.effects(mod_joint_cog_spline)['s.MCI'] * MCI +
                                           fixed.effects(mod_joint_cog_spline)['s.DEM'] * DEM +
                                           fixed.effects(mod_joint_cog_spline)['s.Age_bl'] * Age_bl +
                                           random.effects(mod_joint_cog_spline)[as.character(RID), 's.(Intercept)'])


prog_dat$disease_time_cog_spline_AgeK <- with(prog_dat, Month_bl + 
                                           fixed.effects(mod_joint_cog_spline_AgeK)['s.PRE'] * PRE +
                                           fixed.effects(mod_joint_cog_spline_AgeK)['s.MCI'] * MCI +
                                           fixed.effects(mod_joint_cog_spline_AgeK)['s.DEM'] * DEM +
                                           fixed.effects(mod_joint_cog_spline_AgeK)['s.Age_bl'] * Age_bl +
                                           random.effects(mod_joint_cog_spline_AgeK)[as.character(RID), 's.(Intercept)'])


ggplot(prog_dat, aes(x = disease_time_cog, y = value, group = RID)) + 
  geom_line() +
  facet_wrap(~ name, scales = 'free_y')

ggplot(prog_dat, aes(x = disease_time_cog_spline, y = value, group = RID)) + 
  geom_line() +
  facet_wrap(~ name, scales = 'free_y')

ggplot(prog_dat, aes(x = disease_time_cog_spline_AgeK, y = value, group = RID)) + 
  geom_line() +
  facet_wrap(~ name, scales = 'free_y')
```

# Add Amyloid PET (centiloid)
```{r}
fixed_start_coef <- c(a1_adas = 20, a2_adas = 12, a3_adas = 12, a4_adas = 31, a5_adas = 78, 
                      a1_cdrsb = 0, a2_cdrsb = 0.4, a3_cdrsb = 0.67, a4_cdrsb = 5.3, a5_cdrsb = 18, 
                      a1_mmse = 34, a2_mmse = 29, a3_mmse = 28, a4_mmse = 22, a5_mmse = 4, 
                      a1_AB_centiloid = 1, a2_AB_centiloid = 1.3, a3_AB_centiloid = 1.37, a4_AB_centiloid = 1.4, a5_AB_centiloid = 1.5, 
                      s.PRE = 17, s.MCI = 63, s.DEM = 116, s.Age_bl = 0, k.Age_bl = 0)


random_start <- random.effects(mod_joint_cog_spline)
random_start[, 5] <- 0
names(random_start)[5] <- c('k.nameAB_centiloid')


mod_joint_AB_centiloid <- nlme(value ~ as.numeric(name == 'ADAS13') * spline_mean(Month_bl, s, a1_adas, a2_adas, a3_adas, a4_adas, a5_adas) +
                    as.numeric(name == 'CDRSB') * spline_mean(Month_bl, s, a1_cdrsb, a2_cdrsb, a3_cdrsb, a4_cdrsb, a5_cdrsb) +
                    as.numeric(name == 'MMSE') * spline_mean(Month_bl, s, a1_mmse, a2_mmse, a3_mmse, a4_mmse, a5_mmse) +
                    as.numeric(name == 'AB_centiloid') * spline_mean(Month_bl, s, a1_AB_centiloid, a2_AB_centiloid, a3_AB_centiloid, a4_AB_centiloid, a5_AB_centiloid) + k,
                  data = subset(prog_dat, name %in% c('ADAS13', 'CDRSB', 'MMSE','AB_centiloid')),
                  fixed = list(a1_adas + a2_adas + a3_adas + a4_adas + a5_adas ~ 1,
                               a1_cdrsb + a2_cdrsb + a3_cdrsb + a4_cdrsb + a5_cdrsb ~ 1,
                               a1_mmse + a2_mmse + a3_mmse + a4_mmse + a5_mmse ~ 1,
                               a1_AB_centiloid + a2_AB_centiloid + a3_AB_centiloid + a4_AB_centiloid + a5_AB_centiloid ~ 1,
                               s ~ PRE + MCI + DEM + Age_bl + 0,
                               k ~ Age_bl + 0),
                  random = pdDiag(list(s ~ 1,
                                       k ~ name + 0)),
                  groups = ~ RID,
                  start = list(fixed = fixed_start_coef,
                               random = random_start),
                  weights = varIdent(form = ~ 1 | name),
                  control = nlmeControl(returnObject = TRUE,
                                        msVerbose = TRUE))

AIC(mod_joint_AB_centiloid) # 104187.7
fixed.effects(mod_joint_AB_centiloid)

system("say Amyloid PET added")
```

Update model
```{r, eval=FALSE}
fixed_start_coef <- fixed.effects(mod_joint_AB_centiloid)
random_start <- random.effects(mod_joint_AB_centiloid)

mod_joint_AB_centiloid_updated <- update(mod_joint_AB_centiloid,
                    random = list(s ~ 1,
                                  k ~ name + 0))

# system("say model updated")

orig_model_backup <- mod_joint_AB_centiloid
mod_joint_AB_centiloid <- mod_joint_AB_centiloid_updated

AIC(orig_model_backup, mod_joint_AB_centiloid_updated) # 103573.3
```

```{r, eval=FALSE}
fixed_start_coef <- fixed.effects(mod_joint_AB_centiloid)
random_start <- random.effects(mod_joint_AB_centiloid)

mod_joint_AB_centiloid_updated_2 <- update(orig_model_backup,
                    random = pdBlocked(list(s ~ 1,
                                  k ~ name + 0)))

system("say model updated")

AIC(orig_model_backup, mod_joint_AB_centiloid_updated,mod_joint_AB_centiloid_updated_2)
```

Plot graphs: 
```{r}
prog_dat$disease_time_AB_centiloid <- with(prog_dat, Month_bl +
                              fixed.effects(mod_joint_AB_centiloid)['s.PRE'] * PRE +
                                fixed.effects(mod_joint_AB_centiloid)['s.MCI'] * MCI +
                                fixed.effects(mod_joint_AB_centiloid)['s.DEM'] * DEM +
                                fixed.effects(mod_joint_AB_centiloid)['s.Age_bl'] * Age_bl +
                                random.effects(mod_joint_AB_centiloid)[as.character(RID), 's.(Intercept)'])

ggplot(prog_dat, aes(x = disease_time_AB_centiloid, y= value, group = RID)) + 
  geom_line() +
  facet_wrap(~ name, scales = 'free_y')

m <- 251
mean_dat_AB_centiloid <- data.frame(Month_bl = seq(-100, 150, length = m),
                       PRE = 0,
                       MCI = 0,
                       DEM = 0,
                       Age_bl = 0,
                       name = rep(levels(prog_dat$name), each = m))

mean_dat_AB_centiloid$value <- predict(mod_joint_AB_centiloid, newdata = mean_dat_AB_centiloid, level = 0)

ggplot(prog_dat, aes(x = disease_time_AB_centiloid, y = value, group = RID)) +
  geom_line(alpha=0.3) +
  geom_line(aes(x = Month_bl, y = value), data = mean_dat_AB_centiloid, inherit.aes = FALSE, color = 'red', size = 2) +
  facet_wrap(~ name, scales = 'free_y')

# align_at <- 0
align_at <- min(mean_dat_AB_centiloid$Month_bl)

mean_dat_AB_centiloid <- mean_dat_AB_centiloid %>%
  group_by(name) %>%
  # mutate(value_z = (value-value[Years_since_bl==align_at])/sd(value))
  mutate(value_z = (value-value[Month_bl==align_at])/sd(value))

mean_dat_AB_centiloid$value_z[mean_dat_AB_centiloid$name=='MMSE'] <- mean_dat_AB_centiloid$value_z[mean_dat_AB_centiloid$name=='MMSE']*-1

ggplot(aes(x = Month_bl, y = value_z, group = name, color = name), data = mean_dat_AB_centiloid) +
  # geom_line(aes(x = disease_time, y = value_z, group = RID)) +
  geom_line(size = 2) 
```


Save workspace
```{r, eval=FALSE}
save.image(file="../data/Trajectory_models_covars.Rdata")

save(prog_dat,
     mod_joint_cog, mod_joint_cog_spline, mod_joint_AB_centiloid,
     file="../data/Trajectory_models_essentials_covars.Rdata")
```

```{r, eval=FALSE}
# load(file="../data/Trajectory_models_covars.Rdata")
```

# Save data for publication figures
```{r, eval=FALSE}
save(prog_dat, mod_joint_AB_centiloid, file = "../Publication figures/data/full_trajectory_models.Rdata")
```

