---
title: "Explore_best_Tx_var"
author: "Alfie Wearn"
date: "11/05/2022"
output: pdf_document
---

What is the texture variable with the biggest difference between healthy and MCI groups?

Is there a variable with a big difference between Healthy and Preclinical groups?

Plot these.


```{r}
library(tidyverse)
library(lme4)
library(lmerTest)

remove_outliers <- function(x, na.rm = TRUE, ...) {
  qnt <- quantile(x, probs=c(.25, .75), na.rm = na.rm, ...)
  H <- 1.5 * IQR(x, na.rm = na.rm)
  y <- x
  y[x < (qnt[1] - H)] <- NA
  y[x > (qnt[2] + H)] <- NA
  y
}
```


```{r}
load(file="../data/pc_dataset_4Dx.Rdata")

# Names of TA vars
load(file="../data/TA_names_avg.Rdata")
```

# Filter data
Take only those with Amyloid status - PET tau? No, not very many of those.
```{r}
pc_dataset$DX_TA_baseline <- factor(pc_dataset$DX_TA_baseline, levels = c("CN","MCI","Dementia"))
pc_dataset$PTAU_status <- factor(pc_dataset$PTAU_status, levels = c(0, 1))
pc_dataset$ABETA_status <- factor(pc_dataset$ABETA_status, levels = c(0, 1))

bl_dataset <- pc_dataset %>%
  filter(!is.na(DX_TA_baseline_all)) %>% 
  filter(Month_TA_bl-Month==0) %>%
  filter(!is.na(PTAU_ABETA_status)) %>%
  group_by(RID) %>%
  mutate(Age_TA_bl = min(Age), .after = Age) %>%
  ungroup()
```

Create function to loop through.
```{r}
var <- v
out_list <- results


compare_bmks <- function(var, out_list = NULL, loop_count = NULL) {
  
  print(var)
  print(loop_count)
  
  var_data <- bl_dataset %>%
    select(RID, M, all_of(var), DX_TA_baseline_all, ABETA_status, PTAU_status, PTAU_ABETA_status, Age_TA_bl, male, PTEDUCAT, ICV_bl, SITE)
  var_filt <- var_data %>% select(all_of(var))
  colnames(var_filt) <- 'var_filt'
  var_data <- cbind(var_data, var_filt)

  colnames(var_data)[3] <- 'var'
    
  var_data$SITE <- as.factor(var_data$SITE)
  var_data$male <- as.factor(var_data$male)
  var_data$Age_TA_bl <- scale(var_data$Age_TA_bl)
  var_data$PTEDUCAT <- scale(var_data$PTEDUCAT)
  var_data$ICV_bl <- scale(var_data$ICV_bl)
  var_data$var_filt <- scale(var_data$var_filt)

  # cn_data <-  var_data[var_data$DX_TA_baseline_all=="CN",]
  # cn_data$var_filt <- remove_outliers(cn_data$var_filt)
  # cn.aov <-  lmer(var_filt ~ ABETA_status + Age_TA_bl + male + PTEDUCAT + ICV_bl + (1 | SITE), data = cn_data)

  # prodromal_data <- var_data[var_data$DX_TA_baseline_all!="Dementia",]
  prodromal_data <- var_data[var_data$DX_TA_baseline_all %in% c('Healthy', 'Preclinical'),]
  prodromal_data$var_filt <- remove_outliers(prodromal_data$var_filt)
  prodromal.aov <-  lmer(var_filt ~ DX_TA_baseline_all + PTAU_ABETA_status + Age_TA_bl + male + PTEDUCAT + ICV_bl + (1 | SITE), data = prodromal_data)

    if (is.null(out_list)){
      output <- list()
      output[[1]] <- anova(prodromal.aov)
      names(output)[1:3] <- var

    } else {
      output <- out_list
      output[[loop_count]] <- anova(prodromal.aov)
      names(output)[loop_count] <- var

    }
    return(output)
        
}
```

Remove columns with very very small (or 0) variance
```{r}
bl_dataset_TAvars <- bl_dataset %>% select(all_of(TA_names_avg))

variances <- bl_dataset_TAvars %>% sapply(function(x) var(x, na.rm = TRUE)) %>% as.numeric()
variance_not0 <- bl_dataset_TAvars[which(variances>1e-5)] %>% names

  
```



Loop through all variables, and save model results for each
```{r}
loop_count <- 1
results <- list() #create empty list to contain all data
for (v in variance_not0) {
  
  results <- compare_bmks(v, results, loop_count)
  loop_count <- loop_count + 1
  
}
```

Get all F-values for Amyloid effect and plot them
```{r}
amyloid_F <- sapply(results, function(x){as.numeric(x[['F value']][1])})
qnt_99 <- quantile(amyloid_F, probs = 0.99)

var_99qnt <- amyloid_F[amyloid_F>qnt_99] %>% sort
var_99qnt
plot(amyloid_F)
```

Plot result:
```{r}
bl_dataset %>%
  filter(!is.na(DX_TA_baseline_all)) %>%
  filter(Month==Month_TA_bl) %>%
  ggplot(aes(x=fo_Quartile_25_AvgEC, group = DX_TA_baseline_all)) +
  geom_density(aes(fill=DX_TA_baseline_all), alpha = 0.3)

bl_dataset %>%
  filter(!is.na(DX_TA_baseline_all)) %>%
  filter(Month==Month_TA_bl) %>%
  ggplot(aes(x=fo_Low_notch_AvgEC, group = DX_TA_baseline_all)) +
  geom_density(aes(fill=DX_TA_baseline_all), alpha = 0.3)

bl_dataset %>%
  filter(!is.na(DX_TA_baseline_all)) %>%
  filter(Month==Month_TA_bl) %>%
  ggplot(aes(x=glcm_Inv_autocorrelation_e_AvgHipp_32, group = DX_TA_baseline_all)) +
  geom_density(aes(fill=DX_TA_baseline_all), alpha = 0.3)
```


```{r}
# 
# # Scale binwidth to range of data (roughly)
# range_x <- range(plot_data_ph_pc1$var_filt, na.rm=TRUE) %>% diff
# bw <- range_x / 60
# 
# plot_data_ph_pc1 %>%
#     filter((DX_TA_baseline=="Dementia" & ABETA_status==0)==FALSE) %>%
#     ggplot(mapping = aes(x = ABETA_status,
#                          y = var_filt)) +
#     geom_half_violin(aes(fill = ABETA_status), alpha = 0.4, width=0.8, position=position_dodge(0.8)) +
#     geom_half_dotplot(aes(color=ABETA_status, fill = ABETA_status), binaxis = "y", 
#                       binwidth = bw,
#                       dotsize = 0.5,
#                       stackdir="up", width=0.8) +
#     geom_half_boxplot(aes(fill = ABETA_status), width = 0.2, position = position_dodge(width=0.8), notch=FALSE, side = "l", outlier.alpha=0) +
#     facet_grid(~DX_TA_baseline, 
#                space = "free_x", 
#                scales='free',
#                labeller = as_labeller(
#                c(CN = "Cognitively Normal", MCI = "MCI", Dementia = 'Dementia')),) +
#     theme_classic() +
#     theme(strip.background =element_rect(fill="grey95")) +
#     theme(strip.text = element_text(size=15)) +
#     theme(panel.spacing = unit(0.1, "lines")) +
#     theme(legend.position="none") +
#     theme(axis.text = element_text(size = 15), 
#           axis.title = element_text(size = 15), 
#           axis.title.x = element_text(margin = margin(t = 10, r=0, b = 0, l = 0))) +
#     theme(plot.caption = element_text(hjust=0, size = 10)) +
#     labs(title="", 
#          caption="", 
#          x="", 
#          y="Texture (PC1)") +
#     scale_x_discrete(labels=c("0" = "Abeta-", "1" = "Abeta+"))  +
#     add_pvalue(p_values_ph_pc1,
#              label = "p",
#              xmin = "group1",
#              xmax = "group2",
#              y.position = "y_position",
#              label.size = 4) + 
#     ylim(c(-2.5,3))
```

