---
title: "Independent_variance"
author: "Alfie Wearn"
date: "25/05/2022"
output: pdf_document
---

```{r}
library(tidyverse)
library(boot) # Bootstrapping
library(ggpubr) # Panelling
library(ggprism) # For adding p-value
library(effectsize)
```

# Load TA data with PCs
```{r}
getwd()
load(file="../data/pc_dataset.Rdata")

pc_vars <- c('all_Hipp_PC1','all_Hipp_PC2','all_Hipp_PC3','all_Hipp_PC4','all_Hipp_PC5')

```

# Predicting Cognitive Change
```{r}
data <- pc_dataset %>%
  select(PTID, RID, VISCODE, M, Month_TxBl, Month, DX_TxBl,
         Age, PTEDUCAT, male, ICV_bl,
         ADAS13, CDRSB, MMSE,
         all_of(pc_vars),
         Volume_AvgHipp) %>%
  group_by(RID) %>%
  mutate(Age_bl = Age[Month_TxBl==M]) %>%
  mutate(ADAS13_bl = ADAS13[Month_TxBl==M]) %>%
  mutate(CDRSB_bl = CDRSB[Month_TxBl==M]) %>%
  mutate(MMSE_bl = MMSE[Month_TxBl==M]) %>%
  mutate(Volume_AvgHipp_bl = Volume_AvgHipp[Month_TxBl==M]) %>%
  mutate(all_Hipp_PC1_bl = all_Hipp_PC1[Month_TxBl==M]) %>%
  mutate(all_Hipp_PC2_bl = all_Hipp_PC2[Month_TxBl==M]) %>%
  mutate(all_Hipp_PC3_bl = all_Hipp_PC3[Month_TxBl==M]) %>%
  mutate(all_Hipp_PC4_bl = all_Hipp_PC4[Month_TxBl==M]) %>%
  mutate(all_Hipp_PC5_bl = all_Hipp_PC5[Month_TxBl==M]) %>%
  ungroup() %>%
  filter((!is.na(ADAS13) & !is.na(CDRSB) & !is.na(MMSE) &
            !is.na(all_Hipp_PC1) & 
            !is.na(all_Hipp_PC2) & 
            !is.na(all_Hipp_PC3) & 
            !is.na(all_Hipp_PC4) & 
            !is.na(all_Hipp_PC5) &
            !is.na(Volume_AvgHipp) &
            !is.na(ADAS13_bl) & !is.na(CDRSB_bl) & !is.na(MMSE_bl) &
            !is.na(all_Hipp_PC1_bl) &
            !is.na(all_Hipp_PC2_bl) &
            !is.na(all_Hipp_PC3_bl) &
            !is.na(all_Hipp_PC4_bl) &
            !is.na(all_Hipp_PC5_bl) &
            !is.na(Volume_AvgHipp_bl)
  )) 

# For predicting cognitive decline
data_fu <- data %>%
  filter(!(DX_TxBl=="Dementia")) %>%
  filter(M==Month_TxBl+24) %>%
  mutate(ADAS13_diff = ADAS13-ADAS13_bl) %>%
  mutate(CDRSB_diff = CDRSB-CDRSB_bl) %>%
  mutate(MMSE_diff = MMSE-MMSE_bl)

# Predicting Decline sample size
data_fu %>%
  group_by(DX_TxBl) %>%
  summarise(n=n(),
            Mage = mean(Age_bl),
            SDage = sd(Age_bl))

# For predicting age
data_bl <- data %>%
  filter((DX_TxBl=="CN")) %>%
  filter(M==Month_TxBl)

# Predicting age sample size
data_bl %>%
  summarise(n=n(),
            Mage = mean(Age_bl),
            SDage = sd(Age_bl))

# Save ID lists
pred_cog_change_list <- data_fu %>%
  filter(!is.na(DX_TxBl)) %>% 
  select(PTID, VISCODE) %>%
  mutate(pred_cog_change = "Y")

pred_age_list <- data_bl %>%
  select(PTID, VISCODE) %>%
  mutate(pred_age = "Y")

save(pred_cog_change_list, file="../data/ID_lists/pred_cog_change_list.Rdata")
save(pred_age_list, file="../data/ID_lists/pred_age_list.Rdata")
```

```{r}
lm(ADAS13 ~ ADAS13_bl + Age_bl + PTEDUCAT + male + ICV_bl, data = data_fu)  %>% summary
lm(ADAS13 ~ ADAS13_bl + Age_bl + PTEDUCAT + male + ICV_bl + Volume_AvgHipp_bl, data = data_fu) %>% summary
lm(ADAS13 ~ ADAS13_bl + Age_bl + PTEDUCAT + male + ICV_bl + Volume_AvgHipp_bl + all_Hipp_PC1_bl + all_Hipp_PC2_bl + all_Hipp_PC3_bl + all_Hipp_PC4_bl + all_Hipp_PC5_bl, data = data_fu)  %>% summary
```

```{r}
lm(CDRSB ~ CDRSB_bl + Age_bl + PTEDUCAT + male + ICV_bl, data = data_fu) %>% summary
lm(CDRSB ~ CDRSB_bl + Age_bl + PTEDUCAT + male + ICV_bl + Volume_AvgHipp_bl, data = data_fu) %>% summary
lm(CDRSB ~ CDRSB_bl + Age_bl + PTEDUCAT + male + ICV_bl + Volume_AvgHipp_bl + all_Hipp_PC1_bl + all_Hipp_PC2_bl + all_Hipp_PC3_bl + all_Hipp_PC4_bl + all_Hipp_PC5_bl, data = data_fu) %>% summary
```

```{r}
lm(MMSE ~ MMSE_bl + Age_bl + PTEDUCAT + male + ICV_bl, data = data_fu) %>% summary
lm(MMSE ~ MMSE_bl + Age_bl + PTEDUCAT + male + ICV_bl + Volume_AvgHipp_bl, data = data_fu) %>% summary
lm(MMSE ~ MMSE_bl + Age_bl + PTEDUCAT + male + ICV_bl + Volume_AvgHipp_bl + all_Hipp_PC1_bl + all_Hipp_PC2_bl + all_Hipp_PC3_bl + all_Hipp_PC4_bl + all_Hipp_PC5_bl, data = data_fu) %>% summary
```

# Bootstrapping
Uses 'boot' package: https://www.statmethods.net/advstats/bootstrapping.html 

Define function to extract R2 from the model
```{r}
rsq <- function(formula, data, indices) {
  d <- data[indices,] # allows boot to select sample
  fit <- lm(formula, data=d)
  return(summary(fit)$r.square)
}

plot_results <- function(results) {
  colnames(results) <- c("Covars", "Covars+Volume", "Covars+Volume+Tx")
  
  results_long <- results %>% 
  pivot_longer(cols = colnames(results)) 
  
  df_p_val <- rstatix::t_test(results_long, value ~ name) # Get p-values for t-tests
  print(df_p_val)
  
  # Get y.positions for p-values
  max_val <- max(results_long$value) 
  # min_val <- min(results_long$value)
  range_val <- diff(range(results_long$value))
  ypos1 <- max_val + (range_val*0.06) # Lower one should be max + 6% of the data range
  ypos2 <- max_val + (range_val*0.15) # Upper one should be max + 15% of the data range
  
  return(
    results_long %>%
      ggplot(aes(x=name, y = value)) +
      geom_violin(aes(fill = name), alpha = 0.5) +
      geom_boxplot(width=0.2) +
      theme_classic() +
      scale_fill_brewer(palette = "Dark2") +
      ylab(bquote(''~R^2~'')) +
      xlab("Model Predictors") +
      theme(legend.position = "none") + 
      add_pvalue(df_p_val, 
               y.position = c(ypos2, ypos1, ypos2),
               bracket.shorten = c(0.025, 0, 0.025))
  )
}

# Set seed for replicability
set.seed(123)
```

Bootstrap basic model with 1000 replications
ADAS:
```{r}
# Set seed for replicability
set.seed(123)

results_basic_adas <- boot(data=data_fu, statistic=rsq,
   R=1000, formula=ADAS13 ~ ADAS13_bl + Age_bl + PTEDUCAT + male + ICV_bl)
boot.ci(results_basic_adas, type="bca")

results_vol_adas <- boot(data=data_fu, statistic=rsq,
   R=1000, formula=ADAS13 ~ ADAS13_bl + Age_bl + PTEDUCAT + male + ICV_bl + Volume_AvgHipp_bl)
boot.ci(results_vol_adas, type="bca")

results_vol_tx_adas <- boot(data=data_fu, statistic=rsq,
   R=1000, formula=ADAS13 ~ ADAS13_bl + Age_bl + PTEDUCAT + male + ICV_bl + Volume_AvgHipp_bl + all_Hipp_PC1_bl + all_Hipp_PC2_bl + all_Hipp_PC3_bl + all_Hipp_PC4_bl + all_Hipp_PC5_bl)
boot.ci(results_vol_tx_adas, type="bca")

# Stick all the R2 values together into one dataframe, and plot and compare results
results_adas <- cbind(results_basic_adas$t, results_vol_adas$t, results_vol_tx_adas$t) %>% as.data.frame()

plot_adas <- plot_results(results_adas) + ggtitle("2-year ADAS13 Change")
plot_adas 
```
Repeat for CDRSB:
```{r}
results_basic_cdrsb <- boot(data=data_fu, statistic=rsq,
   R=1000, formula=CDRSB ~ CDRSB_bl + Age_bl + PTEDUCAT + male + ICV_bl)
boot.ci(results_basic_cdrsb, type="bca")

results_vol_cdrsb <- boot(data=data_fu, statistic=rsq,
   R=1000, formula=CDRSB ~ CDRSB_bl + Age_bl + PTEDUCAT + male + ICV_bl + Volume_AvgHipp_bl)
boot.ci(results_vol_cdrsb, type="bca")

results_vol_tx_cdrsb <- boot(data=data_fu, statistic=rsq,
   R=1000, formula=CDRSB ~ CDRSB_bl + Age_bl + PTEDUCAT + male + ICV_bl + Volume_AvgHipp_bl + all_Hipp_PC1_bl + all_Hipp_PC2_bl + all_Hipp_PC3_bl + all_Hipp_PC4_bl + all_Hipp_PC5_bl)
boot.ci(results_vol_tx_cdrsb, type="bca")

results_cdrsb <- cbind(results_basic_cdrsb$t, results_vol_cdrsb$t, results_vol_tx_cdrsb$t) %>% as.data.frame()

plot_cdrsb <- plot_results(results_cdrsb) + ggtitle("2-year CDRSB Change")
plot_cdrsb

t.test(results_cdrsb[,1], results_cdrsb[,2])
t.test(results_cdrsb[,1], results_cdrsb[,3])
t.test(results_cdrsb[,2], results_cdrsb[,3])
```

Repeat for MMSE:
```{r}
results_basic_mmse <- boot(data=data_fu, statistic=rsq,
   R=1000, formula=MMSE ~ MMSE_bl + Age_bl + PTEDUCAT + male + ICV_bl)
boot.ci(results_basic_mmse, type="bca")

results_vol_mmse <- boot(data=data_fu, statistic=rsq,
   R=1000, formula=MMSE ~ MMSE_bl + Age_bl + PTEDUCAT + male + ICV_bl + Volume_AvgHipp_bl)
boot.ci(results_vol_mmse, type="bca")

results_vol_tx_mmse <- boot(data=data_fu, statistic=rsq,
   R=1000, formula=MMSE ~ MMSE_bl + Age_bl + PTEDUCAT + male + ICV_bl + Volume_AvgHipp_bl + all_Hipp_PC1_bl + all_Hipp_PC2_bl + all_Hipp_PC3_bl + all_Hipp_PC4_bl + all_Hipp_PC5_bl)
boot.ci(results_vol_tx_mmse, type="bca")

results_mmse <- cbind(results_basic_mmse$t, results_vol_mmse$t, results_vol_tx_mmse$t) %>% as.data.frame()

plot_mmse <- plot_results(results_mmse) + ggtitle("2-year MMSE Change")
plot_mmse

t.test(results_mmse[,1], results_mmse[,2])
t.test(results_mmse[,1], results_mmse[,3])
t.test(results_mmse[,2], results_mmse[,3])
```

# Calculating difference score
The R2 doesn't change much because baseline predicts so strongly. Try calculating score difference at every time point
ADAS:
```{r, eval=FALSE}
lm(ADAS13_diff ~ Age_bl + PTEDUCAT + male + ICV_bl, data = data_fu) %>% summary
lm(ADAS13_diff ~ Age_bl + PTEDUCAT + male + ICV_bl + Volume_AvgHipp_bl, data = data_fu) %>% summary
lm(ADAS13_diff ~ Age_bl + PTEDUCAT + male + ICV_bl + Volume_AvgHipp_bl + all_Hipp_PC1_bl + all_Hipp_PC2_bl + all_Hipp_PC3_bl + all_Hipp_PC4_bl + all_Hipp_PC5_bl, data = data_fu) %>% summary
```
CDR:
```{r, eval=FALSE}
lm(CDRSB_diff ~ Age_bl + PTEDUCAT + male, data = data_fu) %>% summary
lm(CDRSB_diff ~ Age_bl + PTEDUCAT + male + Volume_AvgHipp_bl, data = data_fu) %>% summary
lm(CDRSB_diff ~ Age_bl + PTEDUCAT + male + Volume_AvgHipp_bl + all_Hipp_PC1_bl + all_Hipp_PC2_bl + all_Hipp_PC3_bl + all_Hipp_PC4_bl + all_Hipp_PC5_bl, data = data_fu) %>% summary
```
MMSE:
```{r, eval=FALSE}
lm(MMSE_diff ~ Age_bl + PTEDUCAT + male, data = data_fu) %>% summary
lm(MMSE_diff ~ Age_bl + PTEDUCAT + male + Volume_AvgHipp_bl, data = data_fu) %>% summary
lm(MMSE_diff ~ Age_bl + PTEDUCAT + male + Volume_AvgHipp_bl + all_Hipp_PC1_bl + all_Hipp_PC2_bl + all_Hipp_PC3_bl + all_Hipp_PC4_bl + all_Hipp_PC5_bl, data = data_fu) %>% summary
```

## Train/Test sets, then compare median absolute errors
```{r, eval=FALSE}
# Then train the model freshly on 80% (training data) + Baseline sessions of test set. Then, Add follow-up timepoints from the test set, and use predict function to predict that. 
set.seed(123)

all_ids <- data_fu$RID %>% unique
samp_id <- sample(all_ids, length(all_ids)*0.8)

data_train <- data_fu %>%
  filter(RID %in% samp_id)

data_test <- data_fu %>%
  filter(!(RID %in% samp_id))

```


Train the mode on the training set
```{r, eval=FALSE}
model <- lm(ADAS13 ~ ADAS13_bl + Age_bl + PTEDUCAT + male + ICV_bl, data = data_train)
data_test$ADAS13_pred <- predict(model,data_test, level = 0)
data_test$ADAS13_error <- data_test$ADAS13_pred - data_test$ADAS13
abs(data_test$ADAS13_error) %>% median(na.rm=TRUE)

model <- lm(ADAS13 ~ ADAS13_bl + Age_bl + PTEDUCAT + male + ICV_bl + Volume_AvgHipp_bl, data = data_train)
data_test$ADAS13_pred <- predict(model,data_test, level = 0)
data_test$ADAS13_error <- data_test$ADAS13_pred - data_test$ADAS13
abs(data_test$ADAS13_error) %>%  median(na.rm=TRUE)

model <- lm(ADAS13 ~ ADAS13_bl + Age_bl + PTEDUCAT + male + ICV_bl + Volume_AvgHipp_bl + all_Hipp_PC1_bl + all_Hipp_PC2_bl + all_Hipp_PC3_bl + all_Hipp_PC4_bl + all_Hipp_PC5_bl, data = data_train)
data_test$ADAS13_pred <- predict(model,data_test, level = 0)
data_test$ADAS13_error <- data_test$ADAS13_pred - data_test$ADAS13
abs(data_test$ADAS13_error) %>%  median(na.rm=TRUE)
```

```{r, eval=FALSE}
model <- lm(CDRSB ~ CDRSB_bl + Age_bl + PTEDUCAT + male + ICV_bl, data = data_train)
data_test$CDRSB_pred <- predict(model,data_test, level = 0)
data_test$CDRSB_error <- data_test$CDRSB_pred - data_test$CDRSB
abs(data_test$CDRSB_error) %>%  median(na.rm=TRUE)

model <- lm(CDRSB ~ CDRSB_bl + Age_bl + PTEDUCAT + male + ICV_bl + Volume_AvgHipp_bl, data = data_train)
data_test$CDRSB_pred <- predict(model,data_test, level = 0)
data_test$CDRSB_error <- data_test$CDRSB_pred - data_test$CDRSB
abs(data_test$CDRSB_error) %>%  median(na.rm=TRUE)

model <- lm(CDRSB ~ CDRSB_bl + Age_bl + PTEDUCAT + male + ICV_bl + Volume_AvgHipp_bl + all_Hipp_PC1_bl + all_Hipp_PC2_bl + all_Hipp_PC3_bl + all_Hipp_PC4_bl + all_Hipp_PC5_bl, data = data_train)
data_test$CDRSB_pred <- predict(model,data_test, level = 0)
data_test$CDRSB_error <- data_test$CDRSB_pred - data_test$CDRSB
abs(data_test$CDRSB_error) %>%  median(na.rm=TRUE)
```

```{r, eval=FALSE}
model <- lm(MMSE ~ MMSE_bl + Age_bl + PTEDUCAT + male + ICV_bl, data = data_train)
data_test$MMSE_pred <- predict(model,data_test, level = 0)
data_test$MMSE_error <- data_test$MMSE_pred - data_test$MMSE
abs(data_test$MMSE_error) %>%  median(na.rm=TRUE)

model <- lm(MMSE ~ MMSE_bl + Age_bl + PTEDUCAT + male + ICV_bl + Volume_AvgHipp_bl, data = data_train)
data_test$MMSE_pred <- predict(model,data_test, level = 0)
data_test$MMSE_error <- data_test$MMSE_pred - data_test$MMSE
abs(data_test$MMSE_error) %>%  median(na.rm=TRUE)

model <- lm(MMSE ~ MMSE_bl + Age_bl + PTEDUCAT + male + ICV_bl + Volume_AvgHipp_bl + all_Hipp_PC1_bl + all_Hipp_PC2_bl + all_Hipp_PC3_bl + all_Hipp_PC4_bl + all_Hipp_PC5_bl, data = data_train)
data_test$MMSE_pred <- predict(model,data_test, level = 0)
data_test$MMSE_error <- data_test$MMSE_pred - data_test$MMSE
abs(data_test$MMSE_error) %>%  median(na.rm=TRUE)
```


# Predicting Age 
```{r}
lm(Age_bl ~ PTEDUCAT + male + ICV_bl, data = data_bl)  %>% summary
lm(Age_bl ~ PTEDUCAT + male + ICV_bl + Volume_AvgHipp_bl, data = data_bl)  %>% summary
lm(Age_bl ~ PTEDUCAT + male + ICV_bl + Volume_AvgHipp_bl + all_Hipp_PC1_bl + all_Hipp_PC2_bl + all_Hipp_PC3_bl + all_Hipp_PC4_bl + all_Hipp_PC5_bl, data = data_bl)  %>% summary

results_basic_age <- boot(data=data_bl, statistic=rsq,
   R=1000, formula=Age_bl ~ PTEDUCAT + male + ICV_bl)

results_vol_age <- boot(data=data_bl, statistic=rsq,
   R=1000, formula=Age_bl ~ PTEDUCAT + male + ICV_bl + Volume_AvgHipp_bl)

results_vol_tx_age <- boot(data=data_bl, statistic=rsq,
   R=1000, formula=Age_bl ~ PTEDUCAT + male + ICV_bl + Volume_AvgHipp_bl + all_Hipp_PC1_bl + all_Hipp_PC2_bl + all_Hipp_PC3_bl + all_Hipp_PC4_bl + all_Hipp_PC5_bl)

results_age <- cbind(results_basic_age$t, results_vol_age$t, results_vol_tx_age$t) %>% as.data.frame()

plot_age <- plot_results(results_age) + ggtitle("Age at baseline")
plot_age

t.test(results_age[,1], results_age[,2])
t.test(results_age[,1], results_age[,3])
t.test(results_age[,2], results_age[,3])
```

Panelling plots
```{r}
ggarrange(plot_adas, plot_cdrsb, plot_mmse, plot_age, labels = c("A","B","C","D"))

ggsave(filename="../figs/R2_Change_Predict_Age_CogDecline.pdf",
       width = 8, 
       height = 8)

```


# Code graveyard
```{r, eval=FALSE}
bl_dataset <- pc_dataset %>%
  filter(!is.na(DX_TxBl)) %>% 
  filter(Month_TxBl-Month==0) %>%
  # filter(M==0) %>%
  # filter(!is.na(PTAU_status)) %>%
  # filter(!is.na(PTAU_ABETA_status)) %>%
  filter(!is.na(ABETA_status)) %>%
  group_by(RID) %>%
  mutate(Age_TA_bl = min(Age), .after = Age) %>%
  ungroup()
```


Do texture an volume explain different variance in predicting: Age, Amyloid Status, or group?
```{r, eval=FALSE}
bl_dataset$SITE <- as.factor(bl_dataset$SITE)
  bl_dataset$male <- as.factor(bl_dataset$male)
  bl_dataset$Age_TA_bl <- scale(bl_dataset$Age_TA_bl)
  bl_dataset$PTEDUCAT <- scale(bl_dataset$PTEDUCAT)
  bl_dataset$ICV_bl <- scale(bl_dataset$ICV_bl)
  bl_dataset$Volume_AvgHipp <- scale(bl_dataset$Volume_AvgHipp)
  
  
  bl_dataset$all_Hipp_PC1 <- remove_outliers(bl_dataset$all_Hipp_PC1)
  bl_dataset$all_Hipp_PC2 <- remove_outliers(bl_dataset$all_Hipp_PC2)
  bl_dataset$all_Hipp_PC3 <- remove_outliers(bl_dataset$all_Hipp_PC3)
  bl_dataset$all_Hipp_PC4 <- remove_outliers(bl_dataset$all_Hipp_PC4)
  bl_dataset$all_Hipp_PC5 <- remove_outliers(bl_dataset$all_Hipp_PC5)
```


```{r, eval=FALSE}
mod_null <-  lm(Age_TA_bl ~ male + PTEDUCAT + ICV_bl, data = bl_dataset)
summary(mod_null)

mod_vol <- lm(Age_TA_bl ~ Volume_AvgHipp + male + PTEDUCAT + ICV_bl, data = bl_dataset)
summary(mod_vol)

mod_vol_h4 <- lm(Age_TA_bl ~  all_Hipp_PC4 + Volume_AvgHipp + male + PTEDUCAT + ICV_bl, data = bl_dataset)
summary(mod_vol_h4)

mod_vol_h1 <- lm(Age_TA_bl ~  all_Hipp_PC1 + Volume_AvgHipp + male + PTEDUCAT + ICV_bl, data = bl_dataset)
summary(mod_vol_h1)

mod_vol_h12 <- lm(Age_TA_bl ~  all_Hipp_PC1 + all_Hipp_PC2 + Volume_AvgHipp + male + PTEDUCAT + ICV_bl, data = bl_dataset)
summary(mod_vol_h12)

mod_vol_h12345 <- lm(Age_TA_bl ~  all_Hipp_PC1 + all_Hipp_PC2 + all_Hipp_PC3 + all_Hipp_PC4 + all_Hipp_PC5 + Volume_AvgHipp + male + PTEDUCAT + ICV_bl, data = bl_dataset)
summary(mod_vol_h12345)

```

