---
title: "Predict_trajectories"
author: "Alfie Wearn"
date: "28/03/2022"
output: pdf_document
---

Load packages and functions
```{r}
library(progmod)
library(tidyverse)
library(splines)
library(lme4)
library(moments) # For calculating skewness
library(RColorBrewer) # For nice plot colours
library(lmerTest)
```

Load data
```{r}
# Calculated disease times
# load(file="../data/Trajectory_models_essentials.Rdata")
load(file="../data/Trajectory_models_essentials_covars.Rdata")

# Names of TA vars
load(file="../data/TA_names.Rdata")

# Load TA data with PCs 
load(file="../data/pc_dataset.Rdata")

```


Choose which texture variables will be predicted
```{r}
# pc_vars <- c('all_x_PC1','all_x_PC2','all_x_PC3','all_x_PC4','all_x_PC5',
#              'fo_x_PC1','fo_x_PC2',
#              'glcm_x_PC1','glcm_x_PC2','glcm_x_PC3','glcm_x_PC4',
#              'glrlm_x_PC1','glrlm_x_PC2','glrlm_x_PC3','glrlm_x_PC4',
#              'Volume_Avgx')

pc_vars <- c('all_x_PC1','all_x_PC2','all_x_PC3','all_x_PC4','all_x_PC5',
             # 'fo_x_PC1','fo_x_PC2',
             # 'glcm_x_PC1','glcm_x_PC2','glcm_x_PC3','glcm_x_PC4',
             # 'glrlm_x_PC1','glrlm_x_PC2','glrlm_x_PC3','glrlm_x_PC4',
             'Volume_Avgx')

# roi <- "AntHipp"
roi_vars_h <- str_replace_all(pc_vars, 'x', 'Hipp')
# roi_vars_ba35 <- str_replace_all(pc_vars, 'x', 'AntHipp')

# Stitch them together
# roi_vars <- c(roi_vars_h,roi_vars_ah,roi_vars_ph,roi_vars_ec)
# roi_vars <- c(roi_vars_h,roi_vars_ah,roi_vars_ph)
roi_vars <- roi_vars_h
```



# Sample size info
Before imputing the missing disease times
```{r}
pc_dataset %>%
  group_by(RID) %>%
  filter(M==min(M)) %>%
  ungroup() %>%
  group_by(DX_baseline_Ab) %>%
  summarise(n=n())

pc_dataset %>%
  group_by(RID) %>%
  filter(M==min(M)) %>%
  ungroup() %>%
  filter(!is.na(all_Hipp_PC1) & (!is.na(CDRSB) | !is.na(MMSE) | !is.na(ADAS13) | !is.na(AB_centiloid))) %>% # Need only rows with all 4 model parameters
  group_by(DX_baseline) %>%
  summarise(n=n(),
            Mage = mean(Age_bl),
            SDage = sd(Age_bl))

```

Explore distributions of ta variables
```{r}
# How many plots to plot? (25, or number available. Whichever is smaller)
num_plots <- min(25, length(roi_vars))

Tx_density_plot_precorrect <- pc_dataset %>%
  filter(!is.na(DX_TxBl_Ab)) %>%
  filter(Month_TxBl==M) %>%
  pivot_longer(roi_vars[1:num_plots], names_to = "measure", values_to = "value") %>%
  ggplot(aes(x=value, group = DX_TxBl_Ab)) +
  facet_wrap(vars(fct_inorder(measure)), scales = "free") +
  geom_density(aes(fill=DX_TxBl_Ab), alpha = 0.3)
Tx_density_plot_precorrect
```

# Transform skewed PCs
```{r, eval=FALSE}
  # Select only the highly skewed columns (absolute skew greater than 1)
  skewness <- pc_dataset %>%
  filter(Month_TxBl==M) %>%
  select(all_of(roi_vars)) %>%
  apply(2, function(x) skewness(x, na.rm=TRUE))
  names_highskew <- which(abs(skewness) > 1) %>% names
  highskew_data <- pc_dataset[,names_highskew]
  lowskew_data <- pc_dataset %>% select(- all_of(names_highskew))
  
  # Log Transform Columns
  # data_plus <- highskew_data %>% sapply(function(x) x-min(x, na.rm=TRUE)+1) %>% as.data.frame 
  # data_xfm <- data_plus %>% sapply(function(x) log(x)) %>% as.data.frame 
  data_plus <- highskew_data - min(highskew_data, na.rm=TRUE) + 1 %>% as.data.frame()
  data_xfm <- log(data_plus) %>% as.data.frame()
  tmp <- cbind(data_xfm, lowskew_data)
  xfm_data <- tmp[,roi_vars]
  
  pc_dataset[,roi_vars] <- xfm_data
  
  Tx_density_plot_xfm <- pc_dataset %>%
  filter(!is.na(DX_TxBl_Ab)) %>%
  filter(Month_TxBl==M) %>%
  pivot_longer(roi_vars[1:num_plots], names_to = "measure", values_to = "value") %>%
  ggplot(aes(x=value, group = DX_TxBl_Ab)) +
  facet_wrap(vars(fct_inorder(measure)), scales = "free") +
  geom_density(aes(fill=DX_TxBl_Ab), alpha = 0.3)
Tx_density_plot_xfm
```


# Exclude outliers
The Trajectory plots are quite sensitive to outliers, so the variables either need to be transformed, or outliers excluded. There are relatively few outlying points, which, in many cases, are very extreme, so excluding them is a better option than transforming the entire dataset.
No need to run this though if the outlier exclusion already occurs in the PCA script. 
```{r, eval=TRUE}
remove_outliers <- function(x, na.rm = TRUE, ...) {
  qnt <- quantile(x, probs=c(.25, .75), na.rm = na.rm, ...)
  H <- 3 * IQR(x, na.rm = na.rm)
  y <- x
  y[x < (qnt[1] - H)] <- NA
  y[x > (qnt[2] + H)] <- NA
  y
}

tmp <- pc_dataset
tmp[roi_vars] <- 
  apply(tmp[roi_vars],2,remove_outliers)

pc_dataset <- tmp

tmp %>%
  filter(!is.na(DX_TxBl_Ab)) %>%
  filter(Month_TxBl==M) %>%
  pivot_longer(roi_vars[1:num_plots], names_to = "measure", values_to = "value") %>%
  ggplot(aes(x=value, group = DX_TxBl_Ab)) +
  facet_wrap(vars(fct_inorder(measure)), scales = "free") +
  geom_density(aes(fill=DX_TxBl_Ab), alpha = 0.3)
```


# Calculate Trajectories
Put the calculated disease times into the main dataset 
```{r}
disease_times <- pivot_wider(prog_dat) %>%
  select(RID, Month_bl, disease_time_AB_centiloid)

progmod_data <-  merge(disease_times, pc_dataset, by = c('RID','Month_bl'), all.y = TRUE)
```

Calculate missing disease times for those with at least one calculated from the model
```{r}
progmod_data <- progmod_data %>%
  group_by(RID) %>%
  filter(any(!is.na(disease_time_AB_centiloid))) %>% # Remove people with no disease times calculated
  mutate(disease_time = disease_time_AB_centiloid[M==min(M)] + Month_bl) %>%
  relocate(disease_time, .after = disease_time_AB_centiloid) %>%
  mutate(age_dt0 = Age_bl-(disease_time_AB_centiloid[M==0]/12)) %>% # Calculate age and disease time 0
  relocate(age_dt0, .after = Age) %>%
  ungroup()
```

# Sample size info
Now that missing disease times have been imputed
```{r}
progmod_data %>%
  group_by(RID) %>%
  filter(M==min(M)) %>%
  ungroup() %>%
  filter(!is.na(disease_time)) %>% 
  group_by(DX_baseline_Ab) %>%
  summarise(n=n())

progmod_data %>%
  group_by(RID) %>%
  filter(M==min(M)) %>%
  ungroup() %>%
  filter((!is.na(all_Hipp_PC1) | !is.na(all_Hipp_PC2) | !is.na(all_Hipp_PC3) | !is.na(all_Hipp_PC4) | !is.na(all_Hipp_PC5)) & (!is.na(CDRSB) | !is.na(MMSE) | !is.na(ADAS13) | !is.na(AB_centiloid))) %>% 
  filter(!is.na(disease_time_AB_centiloid)) %>%
  group_by(DX_baseline_Ab) %>%
  summarise(n=n(),
            Mage = mean(Age_bl),
            SDage = sd(Age_bl))


progmod_data %>%
  group_by(RID) %>%
  filter(M==min(M)) %>%
  filter(!is.na(DX_baseline_Ab)) %>%
  ungroup() %>%
  filter(!is.na(all_Hipp_PC1) & (!is.na(CDRSB) | !is.na(MMSE) | !is.na(ADAS13) | !is.na(AB_centiloid))) %>% # Need only rows with all 4 model parameters
  group_by(DX_baseline) %>%
  summarise(n=n(),
            Mage = mean(Age_bl),
            SDage = sd(Age_bl))

# Save ID list
Tx_model_list <- progmod_data %>%
  filter((!is.na(all_Hipp_PC1) | !is.na(all_Hipp_PC2) | !is.na(all_Hipp_PC3) | !is.na(all_Hipp_PC4) | !is.na(all_Hipp_PC5)) & (!is.na(CDRSB) | !is.na(MMSE) | !is.na(ADAS13) | !is.na(AB_centiloid))) %>% 
  filter(!is.na(disease_time_AB_centiloid)) %>%
  filter(!is.na(DX_baseline_Ab)) %>% 
  select(PTID, VISCODE) %>%
  mutate(Tx_model = "Y")

save(Tx_model_list, file="../data/ID_lists/Tx_model_list.Rdata")
```

Create List of column names to be predicted
```{r}
vars <- c('MMSE', 'CDRSB', 'ADAS13',
          'ABETA','AV1451','AV45','AB_centiloid','PTAU',
          'Hippocampus',
          all_of(roi_vars))
```


Make long dataset with relevant variables
```{r}
cascade_dat <- progmod_data[, c('RID', 
                                'PRE', 'MCI', 'DEM', 'Age','age_dt0',
                                'Month_bl',
                                'disease_time',
                                all_of(vars))]

# Remove v high AV45 value
cascade_dat <- cascade_dat %>%
  filter(RID != 4386) # The one with absurd values for AV45...

# Add in the disease times
# cascade_dat <- merge(cascade_dat, disease_times, by=c('RID','Month_bl'))

# Extend it longer, with all the variables we want to plot (Even if they were not in the original calculation for disease time)
cascade_dat <- pivot_longer(cascade_dat, cols = all_of(vars))
# cascade_dat <- na.omit(cascade_dat) # DOES THIS NEED TO BE COMMENTED OUT?? If this is not commented, each variable will be predicted throughout it's own range. Otherwise, the same disease time range will be used to all variables
cascade_dat$name <- factor(cascade_dat$name)

# cascade_dat <- cascade_dat %>% rename(disease_time = disease_time_AB_centiloid)

```


```{r}
t_p <- seq(-5, 20, length = 300)
cascade_plot <- data.frame(disease_time = NA, name = NA, value = NA, value_d_n = NA)
```

Find variables for which Dementia group is larger than Healthy group. These ones should be 'inverted'
```{r}
healthy_means <- progmod_data %>%
  filter(DX_baseline_Ab=="Healthy") %>%
  select(all_of(vars)) %>%
  colMeans(na.rm=TRUE)

dem_means <- progmod_data %>%
  filter(DX_baseline_Ab=="Dementia") %>%
  select(all_of(vars)) %>%
  colMeans(na.rm=TRUE)

invert_vars <- healthy_means[healthy_means < dem_means] %>% names()
```

## Patient-level sensitivity to change
```{r}
for (v in vars) {
  tmp <- subset(cascade_dat, name == v)
  t_pp <- seq(quantile(tmp$disease_time, p = 0.05), quantile(tmp$disease_time, p = 0.95), length = 200) # Prediction range, excluding 5% limits of data
  mods <- list()
  
  for (df in 3:6) {
    B <- ns(cascade_dat$disease_time, df = df) 
    mods[[df - 2]] <- lmer(value ~ predict(B, newx = disease_time) + (1 | RID), 
                             data = tmp, REML = FALSE)
  }
  
  df <- which.min(sapply(mods, BIC)) + 2
  cat(paste(v, nrow(tmp), ':', round(min(t_pp)), round(max(t_pp)), 'df = ', df, '\n'))
  B <- ns(cascade_dat$disease_time, df = df) 
  mod <- lmer(value ~ predict(B, newx = disease_time) + (1 | RID), 
              data = tmp, REML = TRUE)
  
  direction <- -1
  if (v %in% invert_vars)
    direction <- 1
  
  cascade_plot <- rbind(cascade_plot, 
                        data.frame(
                          disease_time = t_pp, 
                          name = v, 
                          value = predict(mod, 
                                          re.form = ~ 0,
                                          newdata = data.frame(disease_time = t_pp)),
                          value_d_n = direction / summary(mod)$sigma * (predict(mod, 
                                                                                re.form = ~ 0,
                                                                                newdata = data.frame(disease_time = t_pp)) -
                                                                          predict(mod, 
                                                                                  re.form = ~ 0,
                                                                                  newdata = data.frame(disease_time = c(t_pp[1] -  diff(t_pp)[1], t_pp[1:(length(t_pp) - 1)])))) / diff(t_pp)[1]))
}
cascade_plot <- na.omit(cascade_plot)
```



## Mean Scale:
Normalized values based on [0,1] normalization in the central 90% interval of predicted disease times
```{r}
cascade_plot$value_n <- cascade_plot$value

for (v in vars) {
  idx <- cascade_plot$name == v
  if (v %in% invert_vars) {
    cascade_plot$value_n[idx] <- with(cascade_plot[idx, ], (value - min(value)) / diff(range(value)))
  } else {
    cascade_plot$value_n[idx] <- 1 - with(cascade_plot[idx, ], (value - min(value)) / diff(range(value)))
  }
}
```

ALTERNATIVE: Normalized values based on [0,1] normalization in the central 90% interval of predicted disease times BUT not min/max value: min/max predicted time! So always starts and ends in the same place. 
```{r}
# cascade_plot$value_n3 <- cascade_plot$value
# 
# for (v in vars) {
#   idx <- cascade_plot$name == v
#   cascade_plot$value_n3[idx] <- with(cascade_plot[idx, ], (value - value[t_pp==min(disease_time)]) / (value[t_pp==max(disease_time)] - value[t_pp==min(disease_time)]))
# }
```


## Relative abnormality compared to Healthy Group
```{r}
cascade_plot$value_n2 <- cascade_plot$value

for (v in vars) {
  idx <- cascade_plot$name == v
  if (v %in% invert_vars) {
    min_n <- quantile(subset(progmod_data, DX_baseline_Ab=='Healthy')[, v], p = 0.5, na.rm = TRUE)
    max_n <- quantile(subset(progmod_data, DX_baseline_Ab=='Healthy')[, v], p = 0.95, na.rm = TRUE)

    cascade_plot$value_n2[idx] <- with(cascade_plot[idx, ], (value - min_n) / (max_n - min_n))
  } else {
    min_n <- quantile(subset(progmod_data, DX_baseline_Ab=='Healthy')[, v], p = 0.05, na.rm = TRUE)
    max_n <- quantile(subset(progmod_data, DX_baseline_Ab=='Healthy')[, v], p = 0.5, na.rm = TRUE)

    cascade_plot$value_n2[idx] <- with(cascade_plot[idx, ], (value - max_n) / (min_n - max_n))
  }
}
```



# Plots
Plot Prep:
```{r}
# Only keep those variables which go above 0.5 relative to the Healthy group (kind of an arbitrary threshold but it keeps the shape graph looking sensible)
roi_vars_relevant <- cascade_plot %>%
  group_by(name) %>%
  filter((name%in%c(roi_vars)) & value_n2[disease_time==max(disease_time)]>=0.1) %>%
  ungroup()


roi_vars_subset <- roi_vars_relevant$name %>% unique
vol_vars <- roi_vars_subset[grep("Volume_", roi_vars_subset)]
all_TA_vars <- roi_vars_subset[grep("all_", roi_vars_subset)]
roi_vars_subset <- c(vol_vars,all_TA_vars)

plot_data <- cascade_plot %>%    
    filter((name%in%c('ABETA','AB_centiloid',
                      'AV1451','PTAU',
                      'Hippocampus',
                      'CDRSB',
                      all_of(roi_vars_subset))))

# Modify disease time to be the time at which CSF ABETA crosses 95% abnormality relative to CN AB- group
# roi_vars_relevant$disease_time <- 
shift0 <- plot_data %>%
  filter(name == "ABETA" & value_n2 > 0.95)
shift0 <- min(shift0$disease_time)
plot_data$disease_time <- plot_data$disease_time - shift0

# Add columns, such that types of TA/ other vars can be plotted by 'color' and PCs are plotted by line type. 
# Create variable 'groups'...
vars_fo <- roi_vars_subset[grep("fo_",roi_vars_subset)]
vars_glcm <- roi_vars_subset[grep("glcm_",roi_vars_subset)]
vars_glrlm <- roi_vars_subset[grep("glrlm_",roi_vars_subset)]
vars_all <- roi_vars_subset[grep("all_",roi_vars_subset)]

vars_pc1 <- roi_vars_subset[grep("_PC1",roi_vars_subset)]
vars_pc2 <- roi_vars_subset[grep("_PC2",roi_vars_subset)]
vars_pc3 <- roi_vars_subset[grep("_PC3",roi_vars_subset)]
vars_pc4 <- roi_vars_subset[grep("_PC4",roi_vars_subset)]
vars_pc5 <- roi_vars_subset[grep("_PC5",roi_vars_subset)]

vars_vol <- roi_vars_subset[grep("Volume",roi_vars_subset)]
vars_tx <- roi_vars_subset[grep("PC",roi_vars_subset)]

plot_data <- plot_data %>%
  mutate(var_type = ifelse(name %in% vars_fo, "fo", 
                           ifelse(name %in% vars_glcm, "glcm",
                                  ifelse(name %in% vars_glrlm, "glrlm",
                                         ifelse(name %in% vars_all, "All TA",
                                         name))))) %>%
  mutate(pc_type = ifelse(name %in% vars_pc1, "PC1", 
                           ifelse(name %in% vars_pc2, "PC2",
                                  ifelse(name %in% vars_pc3, "PC3",
                                         ifelse(name %in% vars_pc4, "PC4",
                                                ifelse(name %in% vars_pc5, "PC5",
                                                       "Non-texture variable")))))) %>%
  mutate(sub_type = ifelse(name %in% vars_vol, "Volume", 
                           ifelse(name %in% vars_tx, "Texture",
                                  name)))

# plot_data$roi_type <- factor(plot_data$roi_type, levels=c('Non-texture variable','Hipp','AntHipp','PostHipp','EC','BA35'))

cols0 <- colorRampPalette(brewer.pal(5, "PiYG"))
cols1 <- colorRampPalette(brewer.pal(5, "Reds"))
cols2 <- colorRampPalette(brewer.pal(5, "Greens"))
cols3 <- colorRampPalette(brewer.pal(5, "Blues"))
```

Filter which curves should be plotted (optional)
```{r}
# Leave out texture variables
# plot_data <- plot_data[plot_data$name %in% c('ABETA','AB_centiloid', 'AV1451','PTAU', 'Hippocampus', 'CDRSB'),]

# Plot only texture variables
# plot_data <- plot_data[plot_data$name %in% roi_vars_subset,]

# Plot texture variables, amyloid (and a cognitive score?)

plot_data <- plot_data[plot_data$name %in% c('ABETA','CDRSB','PTAU','CDRSB', roi_vars_subset),]

# Reorder these to that order
plot_data$name <- factor(plot_data$name, levels=c('ABETA','CDRSB','PTAU', roi_vars_subset))


```

Rank variables in order of maximum change (relative to healthy group)
```{r}
tmp2 <- cascade_plot %>%
  group_by(name) %>%
  summarise(max_value_n2 = max(value_n2))

abnormality_vars <- tmp2[order(tmp2$max_value_n2, decreasing = TRUE),] 
abnormality_vars
```


Actual plotting
```{r}
# Normalized values on mean scale
plot_data %>%
  ggplot(aes(x=disease_time, y=value_n, group=name, color=sub_type, linetype=pc_type)) +
  geom_line(size = 0.5) +
  # xlim(0,150) +

  # scale_color_manual("",values=c(cols0(1),cols1(2),cols2(1),cols3(4))) +
  ylab("Biomarker abnormality (mean scale)") +
  ggtitle("Shape of variable trajectories")

# # Normalized values on mean scale (allowance for U-shaped trajectory)
# plot_data %>%
#   ggplot(aes(x=disease_time, y=value_n3, group=name, color=var_type, linetype=pc_type)) +
#   geom_line(size = 1) +
#   ylab("Biomarker abnormality (mean scale)") +
#   ggtitle("Shape of variable trajectories (forced start and end and [0,1])")

# Normalized to Healthy group
plot_data %>%
    filter(!(name%in%c('MMSE'))) %>%
  ggplot(aes(x=disease_time, y=value_n2, group=name, color=sub_type, linetype=pc_type)) +
  geom_line(size = 0.5) +
  ylim(-0.15,1.25) +
    # xlim(0,160) +
  ylab("Biomarker abnormality (relative to AB- Healthy Group)") +
  ggtitle("Variable abnormality relative to Healthy (Amyloid-) group")

# Relative Sensitivity
plot_data %>%
  ggplot(aes(x=disease_time, y=value_d_n, group=name, color=sub_type, linetype=pc_type)) +
  geom_line(size = 0.5) +
  # xlim(0,160) +
  ylim(-0.01,0.1) +
  ylab("Relative Sensitivity to Change (patient level)") +
  ggtitle("Shape of variable trajectories")


```
Make these plots prettier
```{r}
plot_data %>%
    filter(!(name%in%c('MMSE'))) %>%
  ggplot(aes(x=disease_time, y=value_n2, group=name, color=sub_type)) +
  geom_line(size = 1.2) +
  # scale_color_manual("",values = c(cols1(4), cols2(2))) +
  # scale_color_brewer(palette = "Dark2") +
# scale_fill_gradientn(colours=coolwarm(100), guide = "colourbar") +
  theme_bw() +  
  ylim(-0.15,1.6) +
  xlim(-24,131) +
  ylab("Biomarker abnormality (relative to AB- Healthy Group)") +
  ggtitle("Variable abnormality relative to Healthy (Amyloid-) group")


plot_data %>%
    filter(!(name%in%c('MMSE'))) %>%
  ggplot(aes(x=disease_time, y=value_d_n, group=name, color=sub_type)) +
  geom_line(size = 1.2) +
  # scale_color_manual("",values = c(cols1(4), cols2(2))) +
  # scale_color_brewer(palette = "Dark2") +
# scale_fill_gradientn(colours=coolwarm(100), guide = "colourbar") +
  theme_bw() +  
  ylim(-0.005, 0.06) +
  xlim(-24,131) +
  ylab("Biomarker abnormality (relative to AB- Healthy Group)") +
  ggtitle("Variable abnormality relative to Healthy (Amyloid-) group")
```





# Explore variables further. 

Relation to age, amyloid status and other covariates
```{r, eval=TRUE}
my_line <- function(x,y,...){
    points(x,y,...)
    fit <- lm(y ~ x) 
    abline(a = fit$coefficients[1] , b = fit$coefficients[2],col="red", lwd=3,...)
}

progmod_data$male <- as.factor(progmod_data$male)

  # Correlation matrix of baseline values
progmod_data %>%
    filter(M==0) %>%
    select(Age, ADAS13, Hippocampus, all_of(roi_vars_subset)) %>%
    pairs(panel = my_line)

# Select a subset of variables to plot trajectory data
# vars_subset <- vars[c(1:14, 25, 26:30,41)]

for (var in c(vars)) {
  # Age by ta_vars
  age_plot <- progmod_data %>%
    ggplot(aes_string(x = "Age", y = var)) +
    geom_line(aes(color=DX_baseline_Ab, group=RID), alpha = 0.5) + 
    geom_point(aes(color=DX_baseline_Ab, group=RID), alpha = 0.2, size = 0.3) + 
    geom_smooth(data = subset(progmod_data, Age > quantile(Age, p = 0.05) & Age < quantile(Age, p = 0.95)), colour="darkgreen") +
    scale_color_brewer(palette = "Dark2") 
  print(age_plot)
  
  dt_plot <- progmod_data %>%
    ggplot(aes_string(x = "disease_time", y = var)) +
    geom_line(aes(color=DX_baseline_Ab, group=RID), alpha = 0.5) + 
    geom_point(aes(color=DX_baseline_Ab, group=RID), alpha = 0.2, size = 0.3) + 
    geom_line(data = subset(cascade_plot, name == var), 
              aes(x= disease_time, y = value), color = 'darkgreen', size = 2) +
    # geom_smooth(data = subset(progmod_data, disease_time> min(t_pp) & disease_time < max(t_pp)), colour="darkgreen") +
    scale_color_brewer(palette = "Dark2") 
  print(dt_plot)
  
  gender_plot <- progmod_data %>%
    subset(M==0) %>%
    ggplot(aes_string(y = var, x= "male")) +
    geom_point(aes(group = male, colour = DX_baseline_Ab), alpha = 0.5, position='jitter') +
    geom_boxplot(aes(group = male), width=0.4, outlier.alpha = 0, alpha = 0.5) +
    scale_color_brewer(palette = "Dark2") 
  print(gender_plot)
  
  group_plot <- progmod_data %>%
    subset(M==0) %>%
    ggplot(aes_string(x = "DX_baseline_Ab", y = var)) +
    geom_point(position = 'jitter') +
    geom_boxplot(width=0.4,  outlier.alpha = 0, alpha = 0.8)
  print(group_plot)
}
```
The lower sensitivity to change of the texture variables could be interpreted as as a result of noise - volume is easier to measure accurately, whereas texture is highly susceptible to noise, despite showing some promise in picking up pathologically relevant information. 


# Save data for publication figures
```{r, eval=FALSE}
save(plot_data, progmod_data, cascade_plot, file = "../Publication figures/data/predict_trajectories.Rdata")
```



# Code Graveyard
## Trajectories age and disease time on the same plot
```{r, eval=FALSE}
for (v in vars) {
  tmp <- subset(cascade_dat, name == v)
  t_pp <- seq(quantile(tmp$disease_time, p = 0.05), quantile(tmp$disease_time, p = 0.95), length = 200) # Prediction range, excluding 5% limits of data
  t_pp_age <- seq(quantile(tmp$age_dt0, p = 0.05), quantile(tmp$age_dt0, p = 0.95), length = 200) # Prediction range, excluding 5% limits of data
  mods <- list()
  for (df in 3:6) {
     B <- ns(cascade_dat$disease_time, df = df) 
     B_age <- ns(cascade_dat$age_dt0, df = df) 
     mods[[df - 2]] <- lmer(value ~ predict(B, newx = disease_time) + predict(B_age, newx = age_dt0) + (1 | RID), 
              data = tmp, REML = TRUE)
  }
  
  df <- which.min(sapply(mods, BIC)) + 2
  cat(paste(v, nrow(tmp), ':', round(min(t_pp)), round(max(t_pp)), 'df = ', df, '\n'))
  B <- ns(cascade_dat$disease_time, df = df) 
  B_age <- ns(cascade_dat$age_dt0, df = df) 
  mod <- lmer(value ~ predict(B, newx = disease_time) + predict(B_age, newx = age_dt0) + (1 | RID), 
              data = tmp, REML = TRUE)
  
  # direction <- -1
  # if (v %in% invert_vars)
  #   direction <- 1
  
  idx <- cascade_plot$name == v
  cascade_plot$value_age[idx]  <- predict(mod,
                                         re.form = ~ 0, 
                                         newdata = data.frame(disease_time = t_pp,
                                                              age_dt0 = t_pp_age))
  
  # cascade_plot$value_age_n[idx] <- with(cascade_plot[idx,], (value_age - min(value_age)) / diff(range(value_age)))
  cascade_plot$age_dt0[idx]  <- t_pp_age
                                                                          
}
```
Normalized values based on [0,1] normalization in the central 90% interval of predicted disease times
```{r, eval=FALSE}
cascade_plot$value_age_n <- cascade_plot$value_age

for (v in vars) {
  idx <- cascade_plot$name == v
  if (v %in% invert_vars) {
    cascade_plot$value_age_n[idx] <- with(cascade_plot[idx, ], (value_age - min(value_age)) / diff(range(value_age)))
  } else {
    cascade_plot$value_age_n[idx] <- 1 - with(cascade_plot[idx, ], (value_age - min(value_age)) / diff(range(value_age)))
  }
}
```

# Age-corrected relative to Healthy Group
```{r, eval=FALSE}
cascade_plot$value_age_n2 <- cascade_plot$value_age

for (v in vars) {
  idx <- cascade_plot$name == v
  if (v %in% invert_vars) {
    min_n <- quantile(subset(progmod_data, DX_baseline_Ab=='Healthy')[, v], p = 0.5, na.rm = TRUE)
    max_n <- quantile(subset(progmod_data, DX_baseline_Ab=='Healthy')[, v], p = 0.95, na.rm = TRUE)

    cascade_plot$value_age_n2[idx] <- with(cascade_plot[idx, ], (value_age - min_n) / (max_n - min_n))
  } else {
    min_n <- quantile(subset(progmod_data, DX_baseline_Ab=='Healthy')[, v], p = 0.05, na.rm = TRUE)
    max_n <- quantile(subset(progmod_data, DX_baseline_Ab=='Healthy')[, v], p = 0.5, na.rm = TRUE)

    cascade_plot$value_age_n2[idx] <- with(cascade_plot[idx, ], (value_age - max_n) / (min_n - max_n))
  }
}
```


Some quick statistical tests
```{r, eval=FALSE}
ta.aov <-  aov(all_Hipp_PC4 ~ DX_baseline_Ab + Age_bl + male + PTEDUCAT + SITE, data = subset(progmod_data,M==0))
summary(ta.aov)

progmod_data %>%
  subset(M==0) %>%
  rstatix::emmeans_test(
    all_Hipp_PC4 ~ DX_baseline_Ab, covariate= c(Age_bl), p.adjust.method = "fdr"
  )
```

LMM change over groups at baseline
```{r, eval=FALSE}
var_data <- progmod_data %>%
  filter(Month_TxBl==M) %>%
  group_by(RID) %>%
  mutate(Age_TA_bl = min(Age), .after = Age) %>%
  ungroup() %>%
  select(RID, M, all_Hipp_PC1, all_Hipp_PC2, all_Hipp_PC3, all_Hipp_PC4, all_Hipp_PC5, Volume_AvgHipp, DX_baseline_Ab, DX_TxBl_Ab, ABETA_status, PTAU_status, PTAU_ABETA_status, Age_TA_bl, male, PTEDUCAT, ICV_bl, SITE)

  var_data$DX_TxBl_Ab <- as.factor(var_data$DX_TxBl_Ab)
  var_data$SITE <- as.factor(var_data$SITE)
  var_data$male <- as.factor(var_data$male)
  var_data$Age_TA_bl <- scale(var_data$Age_TA_bl)
  var_data$PTEDUCAT <- scale(var_data$PTEDUCAT)
  var_data$ICV_bl <- scale(var_data$ICV_bl)
  var_data$all_Hipp_PC4 <- scale(var_data$all_Hipp_PC4)
  var_data$Volume_AvgHipp <- scale(var_data$Volume_AvgHipp)


  # var_data$all_Hipp_PC4 <- remove_outliers(var_data$all_Hipp_PC4)
  
  prodromal_data <- var_data[var_data$DX_TxBl_Ab!="Dementia",]
  prodromal_data <- prodromal_data[prodromal_data$DX_TxBl_Ab!="MCI",]

  
  mod_pc1 <-  lmer(all_Hipp_PC1 ~ DX_TxBl_Ab + Age_TA_bl + male + PTEDUCAT + ICV_bl + (1 | SITE), data = prodromal_data)
  summary(mod_pc1)
  
  mod_pc2 <-  lmer(all_Hipp_PC2 ~ DX_TxBl_Ab + Age_TA_bl + male + PTEDUCAT + ICV_bl + (1 | SITE), data = prodromal_data)
  summary(mod_pc2)
  
  mod_pc3 <-  lmer(all_Hipp_PC3 ~ DX_TxBl_Ab + Age_TA_bl + male + PTEDUCAT + ICV_bl + (1 | SITE), data = prodromal_data)
  summary(mod_pc3)
    
  mod_pc4 <-  lmer(all_Hipp_PC4 ~ DX_TxBl_Ab + Age_TA_bl + male + PTEDUCAT + ICV_bl + (1 | SITE), data = prodromal_data)
  summary(mod_pc4)
  
  mod_pc5 <-  lmer(all_Hipp_PC5 ~ DX_TxBl_Ab + Age_TA_bl + male + PTEDUCAT + ICV_bl + (1 | SITE), data = prodromal_data)
  summary(mod_pc5)
  
  mod_vol <-  lmer(Volume_AvgHipp ~ DX_TxBl_Ab + Age_TA_bl + male + PTEDUCAT + ICV_bl + (1 | SITE), data = prodromal_data)
  summary(mod_vol)
```



```{r, eval=FALSE}
vars
smooth_dat <- subset(progmod_data, disease_time> min(t_pp) & disease_time < max(t_pp))

var = vars[25]
progmod_data %>%
    ggplot(aes_string(x = "disease_time", y = var)) +
    geom_line(aes(color=DX_baseline_Ab, group=RID), alpha = 0.5) + 
    geom_smooth(data = smooth_dat, aes(group=DX_TxBl_Ab, color=DX_TxBl_Ab), method = 'lm') +
    scale_color_brewer(palette = "RdYlBu") 
  
var = vars[14]
progmod_data %>%
    ggplot(aes_string(x = "disease_time", y = var)) +
    geom_line(aes(color=DX_baseline_Ab, group=RID), alpha = 0.5) + 
    geom_smooth(data = smooth_dat, aes(group=DX_TxBl_Ab, color=DX_TxBl_Ab), method = 'lm') +
    scale_color_brewer(palette = "RdYlBu") 
```


