---
title: "TA_PCAs"
author: "Alfie Wearn"
date: "29/03/2022"
output: html_document
---

Create PCAs of glcm variables. Plot out main loadings on each PC (1-5) (we are not excluding any outliers here, should we transform the data?)

```{r}
library(tidyverse)
library(factoextra) # For scree plot
library(moments) # For calculating skewness
library(corrplot)


remove_outliers <- function(x, thresh = 1.5, na.rm = TRUE, ...) {
  qnt <- quantile(x, probs=c(.25, .75), na.rm = na.rm, ...)
  H <- thresh * IQR(x, na.rm = na.rm)
  y <- x
  y[x < (qnt[1] - H)] <- NA
  y[x > (qnt[2] + H)] <- NA
  y
}
```


```{r}
# Dataset with TA vars
load(file="../data/prepped_data.Rdata")
data <- selected_data
# Names of TA vars
load(file="../data/TA_names.Rdata")

```

Get sample size info
```{r}
data %>%
  filter(M==Month_TxBl) %>%
  group_by(DX_TxBl) %>%
  summarise(n=n(),
            Mage = mean(Age_bl),
            SDage = sd(Age_bl))

data %>%
  filter(M==Month_TxBl) %>%
  group_by(RID) %>%
  mutate(Age_TA_bl = min(Age), .after = Age) %>%
  ungroup() %>%
  group_by(DX_TxBl_Ab) %>%
  summarise(n=n(),
            Mage = mean(Age_TA_bl),
            SDage = sd(Age_TA_bl))

# Save ID list
texture_subset_list <- data %>%
  filter(!is.na(DX_TxBl_Ab)) %>%
  select(PTID, VISCODE) %>%
  mutate(texture_subset = "Y")

save(texture_subset_list, file="../data/ID_lists/texture_subset_list.Rdata")

```


```{r}
TA_names_fo <- TA_names[grep("fo_",TA_names)]
TA_names_glcm <- TA_names[grep("glcm_",TA_names)]
TA_names_glrlm <- TA_names[grep("glrlm_",TA_names)]
```

Combine Left and Right Hemispheres
```{r}
# Get Left and Right names separately
TA_names_left <- TA_names[grep("Left",TA_names)]
TA_names_right <- TA_names[grep("Right",TA_names)] # Note: They are not in the same order of regions - this is an embuggerance

# Get the data
Left_TA_data <- data %>% select(all_of(TA_names_left))
Right_TA_data <- data %>% select(all_of(TA_names_right))

# Remove "Left" and "Right' From the names (replace with Avg)
left_names_neutral <- gsub('Left', 'Avg', TA_names_left)
right_names_neutral <- gsub('Right', 'Avg', TA_names_right)

# Rename columns to neutral names
colnames(Left_TA_data) <- left_names_neutral
colnames(Right_TA_data) <- right_names_neutral

# Reorder one dataset so that the column names line up between them
order <- match(left_names_neutral,right_names_neutral)
Left_TA_data_reordered <- Left_TA_data[order(order)]

# Average the two datasets to create Left & Right Average data
Avg_TA_data <- (Left_TA_data_reordered + Right_TA_data) / 2

# Put these average columns into the main dataset
averaged_data <- cbind(data, Avg_TA_data)
```

Get lists for averaged data
```{r}
TA_names_avg <- right_names_neutral[-grep('Volume',right_names_neutral)]
TA_names_avg <- TA_names_avg[-grep('_4',TA_names_avg)]
TA_names_avg <- TA_names_avg[-grep('_8',TA_names_avg)]
TA_names_avg <- TA_names_avg[-grep('_16',TA_names_avg)]

TA_names_avg_fo <- TA_names_avg[grep("fo_",TA_names_avg)]
TA_names_avg_glcm <- TA_names_avg[grep("glcm_",TA_names_avg)]
TA_names_avg_glcm_32 <- TA_names_avg_glcm[grep("_32",TA_names_avg_glcm)]
TA_names_avg_glrlm <- TA_names_avg[grep("glrlm_",TA_names_avg)]
TA_names_avg_glrlm_32 <- TA_names_avg_glrlm[grep("_32",TA_names_avg_glrlm)]

# (Whole) Hipp
TA_names_Hipp <- TA_names_avg[grep('AvgHipp',TA_names_avg)]
TA_names_fo_Hipp <- TA_names_avg_fo[grep('AvgHipp',TA_names_avg_fo)]
TA_names_glcm_Hipp_32 <- TA_names_avg_glcm_32[grep('AvgHipp',TA_names_avg_glcm_32)]
TA_names_glrlm_Hipp_32 <- TA_names_avg_glrlm_32[grep('AvgHipp',TA_names_avg_glrlm_32)]
```

Filter out rows missing vital information
```{r}
vol_cols <- averaged_data %>% colnames() %>% str_detect("Volume") %>% which()
first_vol_col <- vol_cols[1]

averaged_data_nafiltered <- averaged_data %>%
  filter(!is.na(Volume_LeftHipp))

na_per_row <- averaged_data_nafiltered[first_vol_col:length(averaged_data_nafiltered)] %>% apply(1, function(x) sum(is.na(x)))
plot(na_per_row)

# Exclude rows with ANY missing values. If there are none, then great, just take the dataset as-is
if (length(which(na_per_row > 0))) {
  averaged_data_nafiltered <- averaged_data_nafiltered[-which(na_per_row > 0),]
}

na_per_row <- averaged_data_nafiltered[first_vol_col:length(averaged_data_nafiltered)] %>% apply(1, function(x) sum(is.na(x)))
plot(na_per_row)
```


# PCA
Conduct a PCA on Texture data, and Volume data, separately. 
Save the covariate data separately.
```{r}
pca_covars <- averaged_data_nafiltered[1:(first_vol_col+9)]
```

PCA prep function - works out which variables need transforming in which direction, if that's what we're doing
```{r}
list <- TA_names_Hipp

get_skewed_vars <- function(dat, covars, list){
  
  pca_data <- dat %>%
  select(any_of(list))

  # Get baseline healthy data on which to conduct pca
  pca_data_bl <- pca_data %>%
  cbind(covars) %>%
  filter(M==Month_TxBl) %>%
  filter(DX_TxBl=="CN")
  
  covars_bl <- pca_data_bl %>%
  select(any_of(names(covars)))
  
  pca_data_bl <- pca_data_bl %>%
   select(-any_of(names(covars)))
  

  # Select only the highly skewed columns (absolute skew greater than 1)
  skewness <- pca_data_bl %>%
  select(all_of(list)) %>%
  apply(2, function(x) skewness(x))
  names_rightskew <- which(skewness > 1) %>% names # log these
  names_leftskew <- which(skewness < -1) %>% names # Exponent/square these
  names_lowskew <- pca_data_bl %>% select(-all_of(names_rightskew), -all_of(names_leftskew)) %>% names
  

  tmp_r <- replicate(length(names_rightskew), 'R')
  tmp_l <- replicate(length(names_leftskew), 'L')
  tmp_low <- replicate(length(names_lowskew), '0')
  tmp_all <- c(tmp_r, tmp_l, tmp_low)
  names_all <- c(names_rightskew,names_leftskew,names_lowskew)
  
  names(tmp_all) <- names_all
  skewed_vars <- tmp_all[list] %>% unlist() # Order the variables the same as in the original list
  
  return(skewed_vars)
}
```


Define setup pca function
```{r}
setup_pca <- function(dat, covars, list, do_xfm = TRUE, do_skewed_only=FALSE, skewed_vars = NULL, do_exclude_outliers = FALSE) {

  pca_data <- dat %>%
  select(any_of(list))

  # Plot vars before transformation
  num_plots <- min(56, length(list))
  
  plot_before <- dat %>%
  filter(!is.na(DX_TxBl)) %>%
  filter(M==Month_TxBl) %>%
  filter(DX_TxBl=="CN") %>%
  pivot_longer(list[1:num_plots], names_to = "measure", values_to = "value") %>%
  ggplot(aes(x=value, group = DX_TxBl)) +
  facet_wrap(vars(fct_inorder(measure)), scales = "free") +
  geom_density(aes(fill=DX_TxBl), alpha = 0.3) +
    theme_void() +
  ggtitle("Before any changes")
  print(plot_before)
  
    # Get baseline healthy data on which to conduct pca
  pca_data_bl <- pca_data %>%
  cbind(covars) %>%
  filter(M==Month_TxBl) %>%
  filter(DX_TxBl=="CN")
  
  covars_bl <- pca_data_bl %>%
  select(any_of(names(covars)))
  
  pca_data_bl <- pca_data_bl %>%
   select(-any_of(names(covars)))
  
  
  if (do_xfm){
    
    if (do_skewed_only) {
    
    # Require skewed_vars to be calculated elsewhere. This indicates which variables are skewed, and in which direction.
    rightskew_data <- pca_data_bl[,skewed_vars=='R'] %>% as.data.frame()
    leftskew_data <- pca_data_bl[,skewed_vars=='L'] %>% as.data.frame()
    lowskew_data <- pca_data_bl[,skewed_vars=='0'] %>% as.data.frame()
    
    # Have to add names, in case there is only 1 skewed variable in any category
    names(rightskew_data) <- pca_data_bl[skewed_vars=='R'] %>% names
    names(leftskew_data) <- pca_data_bl[skewed_vars=='L'] %>% names
    names(lowskew_data) <- pca_data_bl[skewed_vars=='0'] %>% names
    
    # Log/Exp Transform Columns
    if (length(rightskew_data)==0) {
      data_xfm <- rightskew_data
    } else {
      data_plus <- rightskew_data %>% sapply(function(x) x-min(x, na.rm=TRUE)+1) %>% as.data.frame 
      data_xfm <- data_plus %>% sapply(function(x) log(x)) %>% as.data.frame 
    }
    
      #(Also need to add minimum value so that negative and positive squared values don't get confused)... -5^2 == 5^2 !
    if (length(leftskew_data)==0) {
      data_exp <- leftskew_data
    } else {
      data_exp_plus <- leftskew_data %>% sapply(function(x) x-min(x, na.rm=TRUE)+1) %>% as.data.frame 
      data_exp <- data_exp_plus %>% sapply(function(x) x^2) %>% as.data.frame 
    }
    
    tmp <- cbind(data_xfm, data_exp, lowskew_data)
    pca_data_bl <- tmp[,list]
    
    } else {
    # Log Transform All Columns
    data_plus <- pca_data_bl %>% sapply(function(x) x-min(x, na.rm=TRUE)+1) %>% as.data.frame 
    data_xfm <- data_plus %>% sapply(function(x) log(x)) %>% as.data.frame 
    pca_data_bl <- data_xfm
    }
    
    # Plot vars after transformation
    DX_TxBl <- covars$DX_TxBl
    plot_after <- pca_data_bl %>%
    cbind(covars_bl) %>%
    filter(!is.na(DX_TxBl)) %>%
    filter(M==Month_TxBl) %>%
    filter(DX_TxBl=="CN") %>%
    pivot_longer(list[1:num_plots], names_to = "measure", values_to = "value") %>%
    ggplot(aes(x=value, group = DX_TxBl)) +
    facet_wrap(vars(fct_inorder(measure)), scales = "free") +
    geom_density(aes(fill=DX_TxBl), alpha = 0.3) +
    theme_void() +
    ggtitle("After Transformation")
    print(plot_after)
  }
  

  if (do_exclude_outliers) {

    outliers_excluded <- apply(pca_data_bl,2,function(x) remove_outliers(x, thresh = 3)) %>% as.data.frame()
    
    # I should remove columns that frequenctly cause missing data first...?
    
    
    # Remove rows with missing data
    na_per_row <- outliers_excluded %>% apply(1, function(x) sum(is.na(x)))
    plot(na_per_row)
    
    
    # Exclude rows with ANY missing values - they will cause the PCA to crash. 
    if (length(which(na_per_row > 0))) {
     pca_data_bl <- outliers_excluded[-which(na_per_row > 0),] %>% as.data.frame()
     covars_bl <- covars_bl[-which(na_per_row > 0),] %>% as.data.frame()
    }
    
  }
  
    # Remove columns with a variance of 0 or very very small)
  variances <- pca_data_bl %>% sapply(function(x) var(x, na.rm = TRUE)) %>% as.numeric()
  variance_0 <- list[which(variances<=1e-10)]
  pca_data_bl <- pca_data_bl %>%
    select(-any_of(variance_0))
  
  new_list <- pca_data_bl %>% names()
  
  # Plot data that are to be entered into the PCA.
  plot_after <- pca_data_bl %>%
  cbind(covars_bl) %>%
  filter(!is.na(DX_TxBl)) %>%
  # filter(M==Month_TxBl) %>%
  pivot_longer(new_list[1:(min(num_plots, length(new_list)))], names_to = "measure", values_to = "value") %>%
  ggplot(aes(x=value, group = DX_TxBl)) +
  facet_wrap(vars(fct_inorder(measure)), scales = "free") +
  geom_density(aes(fill=DX_TxBl), alpha = 0.3) +
  theme_void() +
  ggtitle("After Outlier Exclusion")
  print(plot_after)
    
    
  ## PCA
  pca <- prcomp(pca_data_bl, center = TRUE, scale. = TRUE)
  
  # Explore PCA results
  print(fviz_eig(pca, main = new_list[1]))
  print(get_eig(pca))
  
  print(fviz_pca_biplot(pca, repel=TRUE))
  fviz_pca_var(pca)

  # Plot top loadings
  n_load <- 5
  
  loadings <- abs(pca$rotation[,1:6])
  loadings_pc1 <- sort(loadings[,1], decreasing = TRUE)
  loadings_pc1_head <- loadings_pc1 %>% head(n_load)
  loadings_pc1_tail <- loadings_pc1 %>% tail(n_load)
  loadings_pc1_extreme <- c(loadings_pc1_head, loadings_pc1_tail)
  par(mai=c(1,4,1,1))
  print(barplot(loadings_pc1_extreme, horiz=TRUE, las=1))
  # par(mai=c(2,1,1,1))
  # print(barplot(loadings_pc1, las=2, cex.axis = 0.1))

  
  loadings_pc2 <- abs(sort(loadings[,2], decreasing = TRUE))
  loadings_pc2_head <- loadings_pc2 %>% head(n_load)
  loadings_pc2_tail <- loadings_pc2 %>% tail(n_load)
  loadings_pc2_extreme <- c(loadings_pc2_head, loadings_pc2_tail)
  par(mai=c(1,4,1,1))
  print(barplot(loadings_pc2_extreme, horiz=TRUE, las=1))
  
  loadings_pc3 <- abs(sort(loadings[,3], decreasing = TRUE))
  loadings_pc3_head <- loadings_pc3 %>% head(n_load)
  loadings_pc3_tail <- loadings_pc3 %>% tail(n_load)
  loadings_pc3_extreme <- c(loadings_pc3_head, loadings_pc3_tail)
  par(mai=c(1,4,1,1))
  print(barplot(loadings_pc3_extreme, horiz=TRUE, las=1))
  
  loadings_pc4 <- abs(sort(loadings[,4], decreasing = TRUE))
  loadings_pc4_head <- loadings_pc4 %>% head(n_load)
  loadings_pc4_tail <- loadings_pc4 %>% tail(n_load)
  loadings_pc4_extreme <- c(loadings_pc4_head, loadings_pc4_tail)
  par(mai=c(1,4,1,1))
  print(barplot(loadings_pc4_extreme, horiz=TRUE, las=1))
  
  loadings_pc5 <- abs(sort(loadings[,5], decreasing = TRUE))
  loadings_pc5_head <- loadings_pc5 %>% head(n_load)
  loadings_pc5_tail <- loadings_pc5 %>% tail(n_load)
  loadings_pc5_extreme <- c(loadings_pc5_head, loadings_pc5_tail)
  par(mai=c(1,4,1,1))
  print(barplot(loadings_pc5_extreme, horiz=TRUE, las=1))
  
  loadings_pc6 <- abs(sort(loadings[,6], decreasing = TRUE))
  loadings_pc6_head <- loadings_pc6 %>% head(n_load)
  loadings_pc6_tail <- loadings_pc6 %>% tail(n_load)
  loadings_pc6_extreme <- c(loadings_pc6_head, loadings_pc6_tail)
  par(mai=c(1,4,1,1))
  print(barplot(loadings_pc6_extreme, horiz=TRUE, las=1))
  
  return(pca)

}
```

Create function to apply the pca to other rois (choice of pca - TA, fo, glcm, glrlm)
```{r}
apply_PCA_to_roi <- function(pca, dat, covars, roi, do_xfm, do_skewed_only=FALSE, skewed_vars = NULL) {
 
  
  # I will need to remove the column names - I will also need to remove the columns that had 0 variance in the Hipp data!
  column_names <- pca$center %>% names
  column_names_new <- column_names %>% str_replace_all("Hipp", roi)
   
  # Select the new data
  pca_data <- dat %>%
  select(any_of(column_names_new))
  
  # But then change the names to match the old data (so the pca prediction can match them)
  colnames(pca_data) <- column_names
  
  # If variables were transformed before, they need to be transformed here too...
    if (do_xfm){
      if (do_skewed_only) {
        # Require skewed_vars to be calculated elsewhere. This indicates which variables are skewed, and in which direction.
        # Some cols removed for pca due to 0 variance. Need to select only those from skewed_vars list
        skewed_subset <- skewed_vars[column_names]
          
        rightskew_data <- pca_data[,skewed_subset=='R'] %>% as.data.frame()
        leftskew_data <- pca_data[,skewed_subset=='L'] %>% as.data.frame()
        lowskew_data <- pca_data[,skewed_subset=='0'] %>% as.data.frame()
                
        # Have to add names, in case there is only 1 skewed variable in any category
        names(rightskew_data) <- pca_data[skewed_subset=='R'] %>% names
        names(leftskew_data) <- pca_data[skewed_subset=='L'] %>% names
        names(lowskew_data) <- pca_data[skewed_subset=='0'] %>% names
        
        # Log/Exp Transform Columns
        if (length(rightskew_data)==0) {
          data_xfm <- rightskew_data
        } else {
          data_plus <- rightskew_data %>% sapply(function(x) x-min(x, na.rm=TRUE)+1) %>% as.data.frame 
          data_xfm <- data_plus %>% sapply(function(x) log(x)) %>% as.data.frame 
        }
        
        if (length(leftskew_data)==0) {
          data_exp <- leftskew_data
        } else {
          data_exp <- leftskew_data %>% sapply(function(x) x^2) %>% as.data.frame 
        }
        
        tmp <- cbind(data_xfm, data_exp, lowskew_data)
        pca_data <- tmp[,column_names]       
        
        
      } else {
        data_plus <- pca_data %>% sapply(function(x) x-min(x, na.rm=TRUE)+1) %>% as.data.frame 
        data_xfm <- data_plus %>% sapply(function(x) log(x)) %>% as.data.frame 
        pca_data <- data_xfm
      }
    }
  
  
  # Apply PCA transformation to entire dataset 
  pca_all <- predict(pca, pca_data)

  # Save first 5 PCs  
  pc1_to_6 <- pca_all %>% 
    as.data.frame() %>%
    select(PC1:PC6)
  
  # Exclude extreme outliers from this set (I do not know if this is correct...)
  outliers_excluded <- apply(pc1_to_6,2,function(x) remove_outliers(x, thresh=3)) %>% as.data.frame()
  
  pc_plot <- outliers_excluded %>%
    cbind(covars) %>%
    filter(!is.na(DX_TxBl)) %>%
    filter(M==Month_TxBl) %>%
    pivot_longer(1:6, names_to = "measure", values_to = "value") %>%
    # filter(value < 50, value > -50) %>%
    ggplot(aes(x=value, group = DX_TxBl)) +
    facet_wrap(vars(fct_inorder(measure)), scales = "free") +
    geom_density(aes(fill=DX_TxBl), alpha = 0.3) +
    ggtitle(roi)
  
  print(pc_plot)
  
    return(outliers_excluded)
}
```

PCA testing
```{r, eval=FALSE}
dat <- averaged_data_nafiltered
covars <- pca_covars
list <- TA_names_Hipp
do_xfm <-  TRUE
do_exclude_outliers <- TRUE
do_skewed_only <- TRUE
skewed_vars <- get_skewed_vars(averaged_data_nafiltered, pca_covars, TA_names_Hipp) 
# skewed_vars <- skewed_vars_fo

TA_pca <- setup_pca(averaged_data_nafiltered, pca_covars, TA_names_Hipp, do_xfm = TRUE, do_skewed_only=TRUE, skewed_vars = skewed_vars, do_exclude_outliers = TRUE)
TA_pca <- setup_pca(averaged_data_nafiltered, pca_covars, TA_names_Hipp, do_xfm = TRUE, do_skewed_only=TRUE, skewed_vars = skewed_vars, do_exclude_outliers = TRUE)

pca <- TA_pca_all
dat <- averaged_data_nafiltered
covars <- pca_covars
roi <-  "Hipp"

TA_Hipp_all_PCs <- apply_PCA_to_roi(TA_pca_all, averaged_data_nafiltered, pca_covars, roi = "Hipp", do_xfm = do_xfm, do_skewed_only=do_skewed_only, skewed_vars = skewed_vars)

```


# Run the PCAs
Set common arguments
```{r}
dat <- averaged_data_nafiltered
covars <- pca_covars
do_xfm=FALSE
do_skewed_only=FALSE
do_exclude_outliers = TRUE
```

Get skewed vars for each list:
```{r}
skewed_vars <- get_skewed_vars(averaged_data_nafiltered, pca_covars, TA_names_Hipp) 
skewed_vars_fo <- get_skewed_vars(averaged_data_nafiltered, pca_covars, TA_names_fo_Hipp) 
skewed_vars_glcm <- get_skewed_vars(averaged_data_nafiltered, pca_covars, TA_names_glcm_Hipp_32) 
skewed_vars_glrlm <- get_skewed_vars(averaged_data_nafiltered, pca_covars, TA_names_glrlm_Hipp_32) 
```


Setup:
```{r}
TA_pca_all <- setup_pca(dat, covars, TA_names_Hipp, do_xfm = do_xfm, do_exclude_outliers = do_exclude_outliers, do_skewed_only = do_skewed_only, skewed_vars = skewed_vars)
TA_pca_fo <- setup_pca(dat, covars, TA_names_fo_Hipp, do_xfm = do_xfm, do_exclude_outliers = do_exclude_outliers, do_skewed_only = do_skewed_only, skewed_vars = skewed_vars_fo)
TA_pca_glcm <- setup_pca(dat, covars, TA_names_glcm_Hipp_32, do_xfm = do_xfm, do_exclude_outliers = do_exclude_outliers, do_skewed_only = do_skewed_only, skewed_vars = skewed_vars_glcm)
TA_pca_glrlm <- setup_pca(dat, covars, TA_names_glrlm_Hipp_32, do_xfm = do_xfm, do_exclude_outliers = do_exclude_outliers, do_skewed_only = do_skewed_only, skewed_vars = skewed_vars_glrlm)
```



# Apply to all regions
All TA measures:
```{r}
TA_Hipp_all_PCs <- apply_PCA_to_roi(TA_pca_all, averaged_data_nafiltered, pca_covars, roi = "Hipp", do_xfm = do_xfm, do_skewed_only=do_skewed_only, skewed_vars = skewed_vars)
```
FO measures:
```{r}
TA_Hipp_fo_PCs <- apply_PCA_to_roi(TA_pca_fo, averaged_data_nafiltered, pca_covars, roi = "Hipp", do_xfm = do_xfm, do_skewed_only=do_skewed_only, skewed_vars = skewed_vars_fo)
```

GLCM measures:
```{r}
TA_Hipp_glcm_PCs <- apply_PCA_to_roi(TA_pca_glcm, averaged_data_nafiltered, pca_covars, roi = "Hipp", do_xfm = do_xfm, do_skewed_only=do_skewed_only, skewed_vars = skewed_vars_glcm)
```

GLRLM Measures:
```{r}
TA_Hipp_glrlm_PCs <- apply_PCA_to_roi(TA_pca_glrlm, averaged_data_nafiltered, pca_covars, roi = "Hipp", do_xfm = do_xfm, do_skewed_only=do_skewed_only, skewed_vars = skewed_vars_glrlm)
```
# Merge in the PCs to TA data
Put the PCs into the main dataset (at least with RID and M) with new, modality-specific, column names
Careful not to re-run this chunk - it'll keep appending things to the column names!
```{r}
colnames(TA_Hipp_all_PCs) <- paste("all_Hipp", colnames(TA_Hipp_all_PCs), sep = '_')
colnames(TA_Hipp_fo_PCs) <- paste("fo_Hipp", colnames(TA_Hipp_fo_PCs), sep = '_')
colnames(TA_Hipp_glcm_PCs) <- paste("glcm_Hipp", colnames(TA_Hipp_glcm_PCs), sep = '_')
colnames(TA_Hipp_glrlm_PCs) <- paste("glrlm_Hipp", colnames(TA_Hipp_glrlm_PCs), sep = '_')

PC_combined <- cbind(TA_Hipp_all_PCs, TA_Hipp_fo_PCs, TA_Hipp_glcm_PCs, TA_Hipp_glrlm_PCs)

rid_m <- averaged_data_nafiltered %>% select(RID, M)
  
pc_data <- cbind(rid_m, PC_combined)

pc_dataset <- merge(averaged_data, pc_data, by = c('RID','M'), all.x=TRUE)
```


# Save data
```{r}
save(pc_dataset, file="../data/pc_dataset.Rdata")
save(TA_names_avg, file = '../data/TA_names_avg.Rdata')
save(TA_pca_all, file = "../Publication figures/data/TA_pca_all.Rdata") # Save to create figure of top loadings
```


# Make the pdf plots pretty
```{r}
plotdat <- TA_Hipp_all_PCs %>%
  cbind(covars) %>%
  filter(!is.na(DX_TxBl)) %>%
  filter(M==Month_TxBl) %>%
  pivot_longer(1:6, names_to = "measure", values_to = "value") %>%
  select(value, measure, DX_TxBl) %>% 
  mutate(roi = 'Hippocampus')
plotdat$measure <- plotdat$measure %>% as.factor()
levels(plotdat$measure) <-  c('PC1','PC2','PC3','PC4','PC5','PC6')

plotdat %>%
  ggplot(aes(x=value, group = DX_TxBl)) +
  geom_density(aes(fill=DX_TxBl), 
               alpha = 0.3,
               size=0.5
               ) +
  facet_grid(cols = vars(fct_inorder(measure)),
             rows = vars(roi), 
             scales = "free",
             switch = "both"
             ) +
  theme_void() +
  ggtitle("Principal Component distributions") +
  theme(plot.title = element_text(hjust = 0.5),
        strip.text.x = element_text(size=12),
        strip.text.y = element_text(size=12),
        ) +
  scale_fill_brewer(palette=2, type = 'qual', name = "Diagnostic Group") 
```
# Correlation of PCs with demographic etc. variables
```{r}
corrplot_data <- pc_dataset %>%
  filter(!is.na(DX_TxBl)) %>%
  filter(M==Month_TxBl) %>%
  select(Age, PTEDUCAT,
         MMSE, ADAS13, CDRSB,
         PTAU, 
         AB_centiloid,
         names(TA_Hipp_all_PCs)[1:6]) 


M <- cor(corrplot_data,
         use="complete.obs",
         method = 'spearman')
P <- cor.mtest(corrplot_data,
         use="complete.obs",
         method = 'spearman')$p

pc_limit <- nrow(M) - 6

M_new <- M[(pc_limit+1):nrow(M), 1:pc_limit]
P_new <- P[(pc_limit+1):nrow(M), 1:pc_limit]

corrplot(M_new,
         p.mat = P_new,
         method = 'ellipse',
         cl.ratio = 0.4,
         tl.col = 'black',

)
```


# Explore what components comprise the PCs
What makes up the components?
```{r}

  loadings <- (TA_pca_all$rotation[,1:6])
  rownames(loadings) <- rownames(loadings) %>% str_remove_all("_AvgHipp")
  rownames(loadings) <- rownames(loadings) %>% str_remove_all("_32")
  loadings_fo <- loadings[grep("fo_",rownames(loadings)),]
  loadings_glcm <- loadings[grep("glcm_",rownames(loadings)),]
  loadings_glcm1 <- loadings_glcm[1:55,]
  loadings_glcm2 <- loadings_glcm[56:132,]
  loadings_glcm3 <- loadings_glcm[133:180,]
  loadings_glcm4 <- loadings_glcm[181:nrow(loadings_glcm),]
  loadings_glrlm <- loadings[grep("glrlm_",rownames(loadings)),]

ulim <- 0.30
llim <- -0.30

# FO stats
cor_fo <- corrplot(t(loadings_fo), 
           is.corr = FALSE,
           tl.cex = 1,
           tl.col = 'black',
           cl.pos = "b",
           cl.ratio = 0.4,
           col.lim = c(llim,ulim))
cor_fo <- recordPlot()
  
  # GLCM stats
  cor_glcm1 <- corrplot(t(loadings_glcm1), 
         is.corr = FALSE,
           tl.cex = 1,
           tl.col = 'black',
           cl.pos = "b",
           cl.ratio = 0.4,
         col.lim = c(llim,ulim))
  cor_glcm1 <- recordPlot()

   cor_glcm2 <- corrplot(t(loadings_glcm2), 
       is.corr = FALSE,
           tl.cex = 1,
           tl.col = 'black',
           cl.pos = "b",
           cl.ratio = 0.4,
       col.lim = c(llim,ulim))
  cor_glcm2 <- recordPlot()

   cor_glcm3 <- corrplot(t(loadings_glcm3), 
       is.corr = FALSE,
           tl.cex = 1,
           tl.col = 'black',
           cl.pos = "b",
           cl.ratio = 0.4,
         col.lim = c(llim,ulim))
  cor_glcm3 <- recordPlot()

   cor_glcm4 <- corrplot(t(loadings_glcm4), 
           is.corr = FALSE,
           tl.cex = 1,
           tl.col = 'black',
           cl.pos = "b",
           cl.ratio = 0.4,
           col.lim = c(llim,ulim))
  cor_glcm4 <- recordPlot()

  # GLRLM stats
   cor_glrlm <- corrplot(t(loadings_glrlm), 
           is.corr = FALSE,
           tl.cex = 1,
           tl.col = 'black',
           cl.pos = "b",
           cl.ratio = 0.4,
           col.lim = c(llim,ulim))

   cor_glrlm <- recordPlot()
```

Save plots
```{r}
w = 1400
h = 700

png("../figs/loadings_3Dx_fo.png", width = w, height = h)
cor_fo
dev.off()

png("../figs/loadings_3Dx_glcm1.png", width = w, height = h)
cor_glcm1
dev.off()

png("../figs/loadings_3Dx_glcm2.png", width = w, height = h)
cor_glcm2
dev.off()

png("../figs/loadings_3Dx_glcm3.png", width = w, height = h)
cor_glcm3
dev.off()

png("../figs/loadings_3Dx_glcm4.png", width = w, height = h)
cor_glcm4
dev.off()

png("../figs/loadings_3Dx_glrlm.png", width = w, height = h)
cor_glrlm
dev.off()
```


```{r}
loadings <- abs(TA_pca_all$rotation[,1:6])
loadings_pc1 <- sort(loadings[,1], decreasing = TRUE)
loadings_pc2 <- sort(loadings[,2], decreasing = TRUE)
loadings_pc3 <- sort(loadings[,3], decreasing = TRUE)
loadings_pc4 <- sort(loadings[,4], decreasing = TRUE)
loadings_pc5 <- sort(loadings[,5], decreasing = TRUE)
loadings_pc6 <- sort(loadings[,6], decreasing = TRUE)

names_pc1 <- loadings_pc1 %>% names %>% as.data.frame 
names_pc2 <- loadings_pc2 %>% names %>% as.data.frame
names_pc3 <- loadings_pc3 %>% names %>% as.data.frame
names_pc4 <- loadings_pc4 %>% names %>% as.data.frame
names_pc5 <- loadings_pc5 %>% names %>% as.data.frame
names_pc6 <- loadings_pc6 %>% names %>% as.data.frame

names_all <- cbind(names_pc1, names_pc2,names_pc3,names_pc4,names_pc5,names_pc6)
names_all <- names_all %>% lapply(gsub, pattern='_AvgHipp', replacement='')
names_all <- names_all %>% lapply(gsub, pattern='_32', replacement='')
names_all <- names_all %>% as.data.frame
colnames(names_all)  <- c("PC1","PC2","PC3","PC4","PC5","PC6")

# Save the csv of top names
write.csv(names_all, file = "../data/PCA_top_loadings.csv", row.names = FALSE)

plot(loadings_pc1)
plot(loadings_pc2)
plot(loadings_pc3)
plot(loadings_pc4)
plot(loadings_pc5)
plot(loadings_pc6)
```

